---
title: "R Notebook"
output: html_notebook
---

# Spotify Project

Authors: [Bernacchia Alessia](https://github.com/AlessiaBernacchia), [Pioda Tommaso](https://github.com/Thetommigun432), [Villani Giacomo](https://github.com/DownToTheGround)



```{r}
library(vip)
library(readr)
library(dplyr)
library(tidymodels)
library(ggplot2)
library(tidymodels)
library(rsample)
library(tidyr)
library(stringr)
library(purrr)
library(caret)
library(kknn)
library(ggcorrplot)
#install.packages("xgboost")
library(parsnip)
library(randomForest)
library(kknn)
library(xgboost)
```
*Functions*:
```{r}
# count NaN for each column
count_na_per_column <- function(data) {
  sapply(data, function(col) sum(is.na(col)))}

count_na_per_column_df <- function(data) {
  na_counts <- sapply(data, function(col) sum(is.na(col)))
  na_counts_df <- data.frame(
    feature = names(na_counts),
    na_count = na_counts
  )
  na_counts_df %>% arrange(desc(na_count))
}

```


## Import the data
```{r}
billboard <- read_csv("billboard.csv", 
                           col_types = cols(
                            url = col_skip(), 
                            song = col_skip(),
                            instance = col_skip()
                            ))
print(head(billboard,5))
```

```{r}
audio_features <- read_csv("audio_features.csv", 
                          col_types = cols( 
                            performer = col_skip(), 
                            song = col_skip(), 
                            spotify_track_preview_url = col_skip(),
                            spotify_track_explicit = col_skip(),
                            spotify_track_album = col_skip(),
                            spotify_track_id = col_skip()
                            ))
print(head(audio_features,5))
```

```{r}
billboard <- na.omit(billboard) # remove all NaN
audio_features <- as_tibble(audio_features) # translate in a modern and more flexible version
audio_features <- na.omit(audio_features) # remove all NaN

```

```{r}
full_data <- left_join(billboard, audio_features, by = "song_id") # merge datasets
```
## Data exploration and cleaning
```{r}
glimpse(full_data)
```

### Missing Values

```{r}
na_counts <- count_na_per_column_df(full_data)
print(na_counts)
```


```{r}
full_data_clean <- full_data
full_data_clean$previous_week_position[is.na(full_data_clean$previous_week_position)] <- 0 # substitute NaN with 0
full_data_clean <- na.omit(full_data_clean) # remove all NaN

nrow(full_data_clean)
ncol(full_data_clean) 

```
### Distributions
Distribution of numeric data

```{r}
numeric_data <- full_data_clean %>% select(where(is.numeric))

numeric_data_long <- numeric_data %>% pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# histograms distribution facet
ggplot(numeric_data_long, aes(x = value)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 25) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +  # 3 columns
  theme_minimal()
#ggsave("0_numeric_histogram_distribution.png", width = 12, height = 16)
```

Distribution of categorical data
(due the fact there are too many unique values, we choose to filter and observe only the first 50 most frequent variables)
```{r}
categorical_data <- full_data_clean %>% select(where(~ !is.numeric(.)))

categorical_data_long <- categorical_data %>% pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# select top 50
categorical_data_top50 <- categorical_data_long %>%
  group_by(variable, value) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(variable) %>%
  slice_max(order_by = count, n = 50)

# barplot distribution facet
ggplot(categorical_data_top50, aes(x = value, y = count)) +
  geom_col(fill = "steelblue", color = "white") + 
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  theme_minimal() +
  theme(#axis.text.x = element_text(angle = 45, hjust = 1))
    axis.text.x = element_blank() )
#ggsave("0_categorical_histogram_distribution.png", width = 12, height = 16)
```
### Cleaning
```{r}
hist(full_data_clean$instrumentalness)
```

```{r}
hist(full_data_clean$time_signature)
```

```{r}
hist(full_data_clean$speechiness)
```


```{r}
hist(full_data_clean$spotify_track_duration_ms)
```
From the distribution of `instrumentalness` and `time signature` they seem to be not significant for our purpose. Then we decide to remove them.
```{r}
full_data_clean <- full_data_clean %>% select(-c(instrumentalness, time_signature))
```
### Outliers
```{r}
# histograms distribution facet
ggplot(numeric_data_long, aes(x = value, y = "")) +
  geom_boxplot(fill = "steelblue") +
  facet_wrap(~ variable, scales = "free", ncol = 3) + 
  theme_minimal()
#ggsave("0_numeric_boxplot_distribution.png", width = 12, height = 16)
```
```{r}
# correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

ggcorrplot(cor_matrix, 
           method = "square",      # o "circle"
           type = "lower",         # solo triangolo inferiore         
           colors = c("red", "white", "blue"),  # come in seaborn
           title = "Matrice di Correlazione",
           ggtheme = ggplot2::theme_minimal())
```


```{r}
# new features creation
billboard_summary <- billboard %>%
  group_by(song_id) %>%
  summarise(
    min_position = min(week_position),
    max_position = max(week_position),
    avg_position = mean(week_position),
    weeks_in_chart = n_distinct(week_id),
    final_position = last(week_position)
  )

full_data_aggregate <- merge(billboard_summary, audio_features, by = "song_id", all = FALSE)
full_data_aggregate <- full_data_aggregate %>% select(-c(instrumentalness, time_signature))

```

```{r}
print(head(full_data_aggregate, 5))
```


```{r}
nrow(billboard)
nrow(billboard_summary)
nrow(audio_features)
nrow(full_data)
nrow(full_data_aggregate)
```

## Classification- predict whether a sogn will be in the top 10
### Preprocessing
```{r}
classification_data <- full_data_aggregate %>% select(-c( song_id, final_position))
print(head(classification_data,5))
```

```{r}
#create the new variable top10: whether a song is in the top 30
classification_data <- classification_data %>%
  filter(!is.na(spotify_track_popularity)) 

classification_data <- classification_data %>%
  mutate(top30 = ifelse(avg_position <= 10, 1, 0))

# encode popularity
classification_data$spotify_track_popularity <- sapply(classification_data$spotify_track_popularity, function(x) {
  if (x <= 25) {
    return("low")
  }
  if (x > 25 & x <= 50) {
    return("medium")
  }
  if (x > 50 & x <= 75) {
    return("high")
  }
    if (x > 75 & x <= 100) {
    return("very high")
  } 
})
```


```{r}
#remove useless features (usefull only)
classification_data <- classification_data %>% select(-c( avg_position, min_position, max_position))
```


```{r}
# Step 1: Rimuovi caratteri indesiderati
classification_data$spotify_genre <- gsub("\\[|\\]|'", "", classification_data$spotify_genre)

# Step 2: Creare una lista di generi separati da virgola
classification_data$spotify_genre <- strsplit(classification_data$spotify_genre, split = ",")

# Step 3: Estrarre solo il primo genere per ogni riga
classification_data$first_genre <- sapply(classification_data$spotify_genre, function(x) x[1])

classification_data$first_genre <- sapply(classification_data$spotify_genre, function(x) {
  if (length(x) == 0) {
    return("others")
  } else {
    return(x[1])
  }
})

# frequency counter
genre_counts <- table(classification_data$first_genre)

# order based in descrescent order the frequency
genre_counts_sorted <- sort(genre_counts, decreasing = TRUE)

# take only first 15 genres
top_15_genres <- names(genre_counts_sorted)[1:15]

# named others as "others"
classification_data$first_genre <- ifelse(
  classification_data$first_genre %in% top_15_genres,
  classification_data$first_genre,
  "others"
)

# verify
genre_counts_after_grouping <- table(classification_data$first_genre)

# drop original spotify_genre column
classification_data <- subset(classification_data, select = -c(spotify_genre))

# ensure the column is upgraded for dummyVars
df <- data.frame(first_genre = classification_data$first_genre)

# create dummy variables (one-hot encoding)
dummy <- dummyVars(~ first_genre, data = df)
final_df <- data.frame(predict(dummy, newdata = df))

# bind dummy variables to the rest of regression_data (without first_genre)
classification_data <- bind_cols(
  classification_data %>% select(-first_genre),
  final_df
)
```


```{r}
#print number of NaN for each column
count_na_per_column_df(classification_data)
```

```{r}
# select only numerical data
numeric_data_clss <- classification_data %>%
  select(where(is.numeric))

# correlation matrix
cor_matrix <- cor(numeric_data_clss, use = "complete.obs")

ggcorrplot(cor_matrix, 
           method = "square",      # or "circle"
           type = "lower",         # only lower triangle        
           colors = c("red", "white", "blue"),  # come in seaborn
           title = "Correlation Matrix",
           ggtheme = ggplot2::theme_minimal())
```

```{r}
hist(classification_data$top30)
```

```{r}
set.seed(42)
classification_data$top30 <- as.factor(classification_data$top30)
spotify_train_test <- classification_data |> initial_split(prop =  0.7)
spotify_train <- training(spotify_train_test); 
spotify_test  <- testing(spotify_train_test)
```

```{r}
rand_forest_spec <- 
  rand_forest(mtry = 5, trees = 200) |>  
  set_engine('randomForest') |> 
  set_mode('classification')

rf_recipe = recipe(top30~.,
                   data=spotify_train)

rf_wf <- workflow() |> 
  add_recipe(rf_recipe) |> 
  add_model(rand_forest_spec)

rf_wf_fit <- rf_wf |> 
  fit(spotify_train)

rf_wf_fit
```

```{r}
spotify_test_rf <- rf_wf_fit |> 
  augment(spotify_test)

cm_rF <- spotify_test_rf |> 
  conf_mat(truth=top30,
           estimate=.pred_class)
cm_rF
```

```{r}
head(summary(cm_rF),4)
```

```{r}
roc_rF<-roc_curve(spotify_test_rf,
          top30,
          .pred_0,event_level = "first")
roc_rF |> autoplot()
```

```{r}
rf_wf_fit |> 
  extract_fit_engine() |> 
  vip()
```

#hypotesis test

```{r}
numeric_data <- regression_data |> select(where(is.numeric))
par(mfrow = c(2,2))  # cambia a seconda del numero di variabili

for (colname in names(numeric_data)) {
  qqnorm(numeric_data[[colname]], main = paste("QQ Plot -", colname))
  qqline(numeric_data[[colname]])
}
```




```{r}
funk_disco_songs <- numeric_data_clss |>
  filter(first_genrefunk == 1 | first_genredisco == 1) |>
  mutate(genre_group = ifelse(first_genrefunk == 1, "funk", "disco")) |>
  select(loudness, genre_group)
```


```{r}
wilcox.test(loudness ~ genre_group, data = funk_disco_songs)

funk_loudness <- funk_disco_songs$loudness[funk_disco_songs$genre_group == "funk"]
disco_loudness <- funk_disco_songs$loudness[funk_disco_songs$genre_group == "disco"]

plot(density(funk_loudness), col = "blue", lwd = 2, main = "Loudness Density: Funk vs Disco", xlab = "Loudness")
lines(density(disco_loudness), col = "red", lwd = 2)

# Aggiungi legenda
legend("topright", legend = c("Funk", "Disco"), col = c("blue", "red"), lwd = 2)
```
```{r}
pop_funk_songs <- numeric_data_clss |>
  filter(first_genrepop == 1 | first_genrefunk == 1) |>
  mutate(genre_group = ifelse(first_genrepop == 1, "pop", "funk")) |>
  select(danceability, genre_group)
```


```{r}
wilcox.test(danceability ~ genre_group, data = pop_funk_songs)

funk_danceability <- pop_funk_songs$danceability[pop_funk_songs$genre_group == "funk"]
pop_danceability <- pop_funk_songs$danceability[pop_funk_songs$genre_group == "pop"]

plot(density(funk_danceability), col = "blue", lwd = 2, main = "danceability Density: Funk vs Pop", xlab = "Loudness")
lines(density(pop_danceability), col = "red", lwd = 2)

# Aggiungi legenda
legend("topright", legend = c("Funk", "Pop"), col = c("blue", "red"), lwd = 2)

```
```{r}
country <- numeric_data_clss |>
  filter(first_genrecountry == 1) |>
  select(energy, first_genrecountry)
```


```{r}

result <- wilcox.test(country$energy, mu = 0.53)

plot(density(country$energy), main = "Energy Density", xlab = "Energy", col = "steelblue")
abline(v = 0.53, col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("Densità Energy", 
                  "mu = 0.53", 
                  paste("p-value =", round(result$p.value, 4))), 
                  col = c("steelblue", "red", NA), 
                  lwd = c(2, 2, NA), 
                  lty = c(1, 2, NA),
                  bty = "n")

```