---
title: "R Notebook"
output: html_notebook
---

# Spotify Project

 In this project we will analyse the evolution of the music, from 1956 until now.

-   First we start to explore whether is possible to predict the popularity of a song from its audio features

-   In second

link to the **dataset**: [jonhthgeek-DataSetSpotify](https://github.com/rfordatascience/tidytuesday/blob/main/data/2021/2021-09-14/readme.md)

**Authors**: [Bernacchia Alessia](https://github.com/AlessiaBernacchia), [Pioda Tommaso](https://github.com/Thetommigun432), [Villani Giacomo](https://github.com/DownToTheGround)

```{r}
n_samples_to_consider = 1000
```


```{r}
library(vip)
library(readr)
library(dplyr)
library(tidymodels)
library(ggplot2)
library(rsample)
library(tidyr)
library(themis)
library(stringr)
library(purrr)
library(caret)
library(kknn)
library(ggcorrplot)
#install.packages("xgboost")
library(parsnip)
library(randomForest)
library(kknn)
library(xgboost)
library(finetune)
library(doParallel)
library(robustHD)
library(parsnip)

# for exploration
library(corrplot)

# for classification
library(pROC)
```

*Functions*:

```{r}
# count NaN for each column
count_na_per_column <- function(data) {
  sapply(data, function(col) sum(is.na(col)))}

count_na_per_column_df <- function(data) {
  na_counts <- sapply(data, function(col) sum(is.na(col)))
  na_counts_df <- data.frame(
    feature = names(na_counts),
    na_count = na_counts
  )
  na_counts_df %>% arrange(desc(na_count))
}

# extract first genre for the spotify genre
extract_first_genre <- function(df, genre_column = "spotify_genre") {
  # check if column exists
  if (!genre_column %in% colnames(df)) {
    stop(paste("La colonna", genre_column, "non esiste nel dataframe"))
  }
  
  # removes useless not required characters
  df[[genre_column]] <- gsub("\\[|\\]|'", "", df[[genre_column]])
  
  # create a list, with div = ,
  genre_list <- strsplit(df[[genre_column]], split = ",")
  
  # extract only the first genre for each line or rename "others"
  df$first_genre <- sapply(genre_list, function(x) {
    x <- trimws(x)  # remove blanks
    if (length(x) == 0 || is.na(x[1]) || x[1] == "") {
      return("others")
    } else {
      return(x[1])
    }
  })
  
  # convert to categorical value
  df$first_genre <- as.factor(df$first_genre)
  
  # output
  return(df)
}

# process and encode the genre column
process_and_encode_genre <- function(df, genre_column = "first_genre", top_n = 15) {
  # check if column exists
  if (!genre_column %in% colnames(df)) {
    stop(paste("La colonna", genre_column, "non esiste nel dataframe"))
  }
  
  # count the frequency
  genre_counts <- table(df[[genre_column]])
  
  # put them in decrescent order
  genre_counts_sorted <- sort(genre_counts, decreasing = TRUE)
  
  # take top N genres
  top_genres <- names(genre_counts_sorted)[1:min(top_n, length(genre_counts_sorted))]
  
  # rename not in top N into 'others'
  df[[genre_column]] <- ifelse(
    df[[genre_column]] %in% top_genres,
    df[[genre_column]],
    "others"
  )
  
  # convert
  df[[genre_column]] <- as.factor(df[[genre_column]])
  
  # remove original column
  if ("spotify_genre" %in% colnames(df)) {
    df <- subset(df, select = -c(spotify_genre))
  }
  
  # create dummy variables (one-hot encoding)
  dummy <- dummyVars(paste("~", genre_column), data = df)
  dummy_df <- data.frame(predict(dummy, newdata = df))
  
  final_df <- bind_cols(
    df %>% select(-all_of(genre_column)),
    dummy_df
  )
  
  # output
  return(final_df)
}

# encode popularity column
encode_spotify_popularity <- function(df, column = "spotify_track_popularity") {
  # check if column exists
  if (!column %in% colnames(df)) {
    stop(paste("La colonna", column, "non esiste nel dataframe"))
  }
  
  categorize_popularity <- function(x) {
    if (is.na(x)) return(NA)
    if (x <= 25) {
      return("low")
    } else if (x <= 50) {
      return("medium")
    } else if (x <= 75) {
      return("high")
    } else if (x <= 100) {
      return("very high")
    } else {
      return(NA)  # out of range
    }
  }
  
  df[[column]] <- sapply(df[[column]], categorize_popularity)
  
  # convert in ordered factor
  df[[column]] <- factor(df[[column]], 
                         levels = c("low", "medium", "high", "very high"), 
                         ordered = TRUE)
  
  # output
  return(df)
}

# plots the confusion matrix
plot_conf_matrix <- function(conf_matrix_object, title = "Confusion Matrix") {
  
  # converts confusion matrix into dataframe
  cm_df <- as.data.frame(conf_matrix_object$table)
  
  # find columns names (Truth, Prediction, Freq)
  colnames(cm_df) <- tolower(colnames(cm_df)) # rende tutto minuscolo
  
  # rename them
  names(cm_df)[names(cm_df) %in% c("reference", "truth")] <- "Actual"
  names(cm_df)[names(cm_df) == "prediction"] <- "Predicted"
  names(cm_df)[names(cm_df) %in% c("freq", "n")] <- "Freq"
  
  # plot
  p <- ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Freq)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = Freq), size = 5, color = "black") +
    scale_fill_gradient(low = "white", high = "steelblue") +
    labs(title = title,
         x = "Predicted Class",
         y = "Actual Class",
         fill = "Count") +
    theme_minimal(base_size = 14) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  return(p)
}
```

## 1.Import the data

```{r}
billboard <- read_csv("billboard.csv", 
                           col_types = cols(
                            url = col_skip(), 
                            song = col_skip(),
                            instance = col_skip()
                            ))
print(head(billboard,5))
```

```{r}
audio_features <- read_csv("audio_features.csv", 
                          col_types = cols( 
                            performer = col_skip(), 
                            song = col_skip(), 
                            spotify_track_preview_url = col_skip(),
                            spotify_track_explicit = col_skip(),
                            spotify_track_album = col_skip(),
                            spotify_track_id = col_skip()
                            ))
print(head(audio_features,5))
```

```{r}
billboard <- na.omit(billboard) # remove all NaN
audio_features <- as_tibble(audio_features) # translate in a modern and more flexible version
audio_features <- na.omit(audio_features) # remove all NaN

```

```{r}
full_data <- left_join(billboard, audio_features, by = "song_id") # merge datasets
```

## 2.Data exploration and cleaning

```{r}
glimpse(full_data)
```

```{r}
summary(full_data)
```

### 2.1 Missing Values

```{r}
na_counts <- count_na_per_column_df(full_data)
print(na_counts)
```

```{r}
full_data_clean <- full_data
full_data_clean$previous_week_position[is.na(full_data_clean$previous_week_position)] <- 0 # substitute NaN with 0
full_data_clean <- na.omit(full_data_clean) # remove all NaN

nrow(full_data_clean)
ncol(full_data_clean) 

```

### 2.2 Distributions

Distribution of numeric data

```{r}
numeric_data <- full_data_clean %>% select(where(is.numeric))

numeric_data_long <- numeric_data %>% pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# histograms distribution facet
ggplot(numeric_data_long, aes(x = value)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 25) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +  # 3 columns
  theme_minimal()
#ggsave("0_numeric_histogram_distribution.png", width = 12, height = 16)
```

Distribution of categorical data (due the fact there are too many unique values, we choose to filter and observe only the first 50 most frequent variables)

```{r}
categorical_data <- full_data_clean %>% select(where(~ !is.numeric(.)))

categorical_data_long <- categorical_data %>% pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# select top 50
categorical_data_top50 <- categorical_data_long %>%
  group_by(variable, value) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(variable) %>%
  slice_max(order_by = count, n = 50)

# barplot distribution facet
ggplot(categorical_data_top50, aes(x = value, y = count)) +
  geom_col(fill = "steelblue", color = "white") + 
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  theme_minimal() +
  theme(#axis.text.x = element_text(angle = 45, hjust = 1))
    axis.text.x = element_blank() )
#ggsave("0_categorical_histogram_distribution.png", width = 12, height = 16)
```

### 2.3 Cleaning

```{r}
hist(full_data_clean$instrumentalness)
```

```{r}
hist(full_data_clean$time_signature)
```

```{r}
hist(full_data_clean$speechiness)
```

```{r}
hist(full_data_clean$spotify_track_duration_ms)
```

From the distribution of `instrumentalness` and `time signature` they seem to be not significant for our purpose. Then we decide to remove them.

```{r}
full_data_clean <- full_data_clean %>% select(-c(instrumentalness, time_signature))
```

### 2.4 Outliers

```{r}
# histograms distribution facet
ggplot(numeric_data_long, aes(x = value, y = "")) +
  geom_boxplot(fill = "steelblue") +
  facet_wrap(~ variable, scales = "free", ncol = 3) + 
  theme_minimal()
#ggsave("0_numeric_boxplot_distribution.png", width = 12, height = 16)
```

### 2.5 Correlations

```{r}
# compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# correlation heatmap
corrplot(cor_matrix, method = "color", 
         type = "upper", tl.col = "black", 
         #number.cex = 0.7, addCoef.col = "black",
         tl.cex = 0.8, col = colorRampPalette(c("purple", "white", "cyan"))(100))
#ggsave("0_numeric_heatmap.png", width = 12, height = 16)
```

## 3. Global Preprocessing
### 3.1 Aggregations

```{r}
# new features creation
billboard_summary <- billboard %>%
  group_by(song_id) %>%
  summarise(
    min_position = min(week_position),
    max_position = max(week_position),
    avg_position = mean(week_position),
    weeks_in_chart = n_distinct(week_id),
    final_position = last(week_position)
  )

full_data_aggregate <- merge(billboard_summary, audio_features, by = "song_id", all = FALSE)
full_data_aggregate <- full_data_aggregate %>% select(-c(instrumentalness, time_signature))

print(head(full_data_aggregate, 5))
```

```{r}
nrow(billboard)
nrow(billboard_summary)
nrow(audio_features)
nrow(full_data)
nrow(full_data_aggregate)
```
### 3.2 Add time columns
```{r}
full_data_clean <- full_data_clean |>
  select(colnames(full_data_clean),song_id,week_id)

full_data_clean$year <- as.numeric(sub(".*/.*/", "", full_data_clean$week_id))

head(full_data_clean, 5)
```

## 4 Main Question - Given the audio features of a new song, in which position it will chart on Billboard?

### 4.1 Regression - predict the position on the billboard

#### Preprocessing

Remove unrelevant features

```{r}
regression_data <- full_data_aggregate %>% select(-c(song_id, min_position, max_position, avg_position, weeks_in_chart, final_position))
print(head(regression_data,5))
```
One hot encoding of the 15 most frequent genre

```{r}
regression_data <- extract_first_genre(regression_data)
regression_data_encoded <- process_and_encode_genre(regression_data, top_n = 15)
set.seed(45)
```
Scaling
```{r}
regression_data_encoded_scaled <- robustHD::standardize(regression_data_encoded |> select(-spotify_track_popularity))

regression_data_encoded_scaled['spotify_track_popularity'] = regression_data$spotify_track_popularity

print(head(regression_data_encoded_scaled, 5))
```

#### Regressions
1. split the dataset
```{r}
set.seed(45)
split <- initial_split(head(regression_data_encoded,n_samples_to_consider), prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)

# resampling for cross validation
cv_folds <- vfold_cv(train_data, v = 5)
```
2. define recipe, models and workflows
```{r}
# recipe
recipe_spec <- recipe(spotify_track_popularity ~ ., data = train_data) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

# models with tuning parameters
task = 'regression'
models <- list(
  rf = rand_forest(mtry = tune(), trees = 500) |>
    set_engine("randomForest") |> set_mode(task),

  xgb = boost_tree(trees = 1000, tree_depth = tune(), learn_rate = tune()) |>
    set_engine("xgboost") |> set_mode(task),

  knn = nearest_neighbor(neighbors = tune()) |>
    set_engine("kknn") |> set_mode(task)
)

# workflows
workflows_list <- purrr::imap(models, function(model, name) {
  workflow() |> add_model(model) |> add_recipe(recipe_spec)
})
```

3. tuning
```{r}
# tune_race_anova
tuned_models <- purrr::imap(workflows_list, function(wf, name) {
  tune_race_anova(
    wf,
    resamples = cv_folds,
    grid = 5,
    metrics = metric_set(rmse),
    control = control_race(verbose_elim = TRUE)
  )})

# extract best values
best_models <- purrr::map(tuned_models, ~ select_best(.x))
```

4. finalize and fits the models to the training set
```{r}
final_models <- purrr::map2(workflows_list, best_models, ~ finalize_workflow(.x, .y) |> fit(data = train_data))
```

5. evaluation on the test set
```{r}
results <- purrr::map(final_models, ~ {
  predict(.x, test_data) |>
    bind_cols(test_data) |>
    metrics(truth = spotify_track_popularity, estimate = .pred)})

# summary RMSE
results_df <- bind_rows(results, .id = "model") |>
  filter(.metric == "rmse") |> arrange(.estimate)

print(results_df)
```

```{r}
all_predictions <- purrr::imap(final_models, ~ {
  predict(.x, test_data) |>
    bind_cols(test_data) |>
    mutate(model = .y)
}) |> bind_rows()

```

```{r}
ggplot(all_predictions, aes(x = spotify_track_popularity, y = .pred)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~model, scales = "free") +
  labs(
    title = "Predictions vs Real Values for Models",
    x = "Real Values (spotify_track_popularity)",
    y = "Predictions"
  ) +
  theme_minimal()
```

features importance
```{r}
# VIP plot list
vip_plots <- list()

# Random Forest
rf_fit <- extract_fit_parsnip(final_models$rf)
if (inherits(rf_fit$fit, "randomForest")) {
  vip_plots$rf <- vip(rf_fit$fit, num_features = 15) + ggtitle("Random Forest - Variable Importance")
}

# XGBoost
xgb_fit <- extract_fit_parsnip(final_models$xgb)
if (inherits(xgb_fit$fit, "xgb.Booster")) {
  vip_plots$xgb <- vip(xgb_fit$fit, num_features = 15) + ggtitle("XGBoost - Variable Importance")
}

# visualize plots
library(patchwork)

# show them
if (length(vip_plots) == 2) {
  vip_plots$rf + vip_plots$xgb
} else {
  # if only one is possible
  vip_plots[[1]]
}
```


```{r}
'''
# Filter and group data for "rock" genre
rock_loudness <- regression_data %>%
'''
  #filter(first_genre == 'album rock') %>%
 ''' pull(loudness)

jazz_loudness <- regression_data %>%'''
  #filter(first_genre == 'dance pop') %>%
  '''pull(loudness)


# Plot density for the "rock" genre
plot(density(rock_loudness), col = "blue", main = "Compare Densities")

# Assuming y is already defined as another dataset for comparison
lines(density(jazz_loudness), col = "red")

# Add legend
legend("topright", legend = c("Rock", "Y"), col = c("blue", "red"), lty = 1)

ks.test(rock_loudness, jazz_loudness)
'''
```

QQ plot
```{r}
numeric_data <- regression_data |> select(where(is.numeric))
par(mfrow = c(2,2))  # cambia a seconda del numero di variabili

for (colname in names(numeric_data)) {
  qqnorm(numeric_data[[colname]], main = paste("QQ Plot -", colname))
  qqline(numeric_data[[colname]])
}
```


### 4.2 Classification - predict which level of popularity
#### Preprocessing
Remove unrelevant features
```{r}
classification_data <- full_data_aggregate |> select(-c( song_id, min_position, max_position, avg_position, weeks_in_chart, final_position))
print(head(classification_data,5))
```

Encode popularity
```{r}
classification_data <- encode_spotify_popularity(classification_data)

classification_data <- classification_data |>
  filter(!is.na(spotify_track_popularity)) 
```

Encode genre
```{r}
classification_data <- extract_first_genre(classification_data)
classification_data <- process_and_encode_genre(classification_data, top_n = 15)

print(head(classification_data,5))
```

Analyse the correlations
```{r}
# select only numerical data
numeric_data_clss <- classification_data |>
  select(where(is.numeric))

# compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# correlation heatmap
corrplot(cor_matrix, method = "color", 
         type = "upper", tl.col = "black", 
         #number.cex = 0.7, addCoef.col = "black",
         tl.cex = 0.8, col = colorRampPalette(c("purple", "white", "cyan"))(100))

#ggsave("4_regression_heatmap.png", width = 12, height = 16)
```

Target Distribution
```{r}
ggplot(classification_data, aes(x = spotify_track_popularity)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Popularity Distribution",
       x = "Spotify Track Popularity",
       y = "Counter") +
  theme_minimal()
```
split
```{r}
set.seed(42)
spotify_train_test <- classification_data |> initial_split(prop =  0.7)
spotify_train <- training(spotify_train_test) 
spotify_test  <- testing(spotify_train_test)
```

### Classification
```{r}
rand_forest_spec <- 
  rand_forest(mtry = 5, trees = 200) |>  
  set_engine('randomForest') |> 
  set_mode('classification')

rf_recipe = recipe(spotify_track_popularity~.,
                   data=spotify_train)

rf_wf <- workflow() |> 
  add_recipe(rf_recipe) |> 
  add_model(rand_forest_spec)

rf_wf_fit <- rf_wf |> 
  fit(spotify_train)

rf_wf_fit
```

```{r}
spotify_test_rf <- rf_wf_fit |> 
  augment(spotify_test)

# confusion matrix
cm_rF <- spotify_test_rf |> 
  conf_mat(truth = spotify_track_popularity,
           estimate = .pred_class)

# visualize
plot_conf_matrix(cm_rF, title = "Confusion Matrix - Classification Random Forest")
```

```{r}
head(summary(cm_rF),4)
```

```{r}
# select(spotify_test_rf, starts_with(".pred_"))

rf_wf_fit |> 
  extract_fit_engine() |> 
  vip()
```

<<<<<<<<<<<TODO>>>>>>>>>>>>>
> Is the music changed over time?

## 5. Taking only the last 25 years of data, <<<<<<<<the model will improve / possibile stimare meglio>>>>>>>>>>>>>?
### 5.1 Preprocessing
Remove unrelevant features
```{r}
data_filtered <- full_data_clean |> select(-c(week_id, week_position, previous_week_position, peak_position, weeks_on_chart))
```

convert milliseconds in minutes
```{r}
data_filtered$spotify_track_duration_min <- data_filtered$spotify_track_duration_ms/60000
data_filtered <- data_filtered |> select(-c(spotify_track_duration_ms))
```

Filter the song after 2000
```{r}
bound_year = 2000

data_filtered_after <- data_filtered[data_filtered$year >= bound_year & data_filtered$year < 2025, ]
data_filtered_before <- data_filtered[data_filtered$year < bound_year, ]
```

encode the most frequent 15 genres for each period
```{r}
data_filtered_after_encoded <- extract_first_genre(data_filtered_after) 
data_filtered_after_encoded <- process_and_encode_genre(data_filtered_after_encoded, top_n = 10)

data_filtered_before_encoded <- extract_first_genre(data_filtered_before) 
data_filtered_before_encoded <- process_and_encode_genre(data_filtered_before_encoded, top_n = 10)
```

remove duplicates
```{r}
data_filtered_after_encoded <- unique(data_filtered_after_encoded)
data_filtered_before_encoded <- unique(data_filtered_before_encoded)
```

```{r}
print(head(data_filtered_after_encoded, 5))
print(head(data_filtered_before_encoded, 5))
```

### 5.2 Regression after 2000
#### Preprocessing
```{r}
regression_data_after <- data_filtered_after_encoded |> select(-c(performer, song_id))
```

Scale the features
<<<<<<<<<<<<<<<<<TODO: vedere se mettere uno unique per ogni anno --> problema riassegnare target>>>>>>>>>>>>>>>>>>>>
```{r}
regression_data_after <- robustHD::standardize(regression_data_after |> select(-c( year, spotify_track_popularity)))
regression_data_after$spotify_track_popularity <- data_filtered_after_encoded$spotify_track_popularity
regression_data_after <- unique(regression_data_after)

#regression_data_after$year <- data_filtered_after_encoded$year
print(head(regression_data_after, 5))
```
#### Regressions
1. split the dataset
```{r}
registerDoParallel(cores = parallel::detectCores() - 1)

set.seed(45)
split <- initial_split(head(regression_data_after, n_samples_to_consider), prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)

# resampling for cross validation
cv_folds <- vfold_cv(train_data, v = 5)
```

2. define recipe, models and workflows
```{r}
# recipe
recipe_spec <- recipe(spotify_track_popularity ~ ., data = train_data) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

# models with tuning parameters
task = 'regression'
models <- list(
  rf = rand_forest(mtry = tune(), trees = 500) |>
    set_engine("randomForest") |> set_mode(task),

  xgb = boost_tree(trees = 1000, tree_depth = tune(), learn_rate = tune()) |>
    set_engine("xgboost") |> set_mode(task),

  knn = nearest_neighbor(neighbors = tune()) |>
    set_engine("kknn") |> set_mode(task)
)

# workflows
workflows_list <- purrr::imap(models, function(model, name) {
  workflow() |> add_model(model) |> add_recipe(recipe_spec)
})
```

3. tuning
```{r}
# tune_race_anova
tuned_models <- purrr::imap(workflows_list, function(wf, name) {
  tune_race_anova(
    wf,
    resamples = cv_folds,
    grid = 5,
    metrics = metric_set(rmse),
    control = control_race(verbose_elim = TRUE)
  )})

# extract best values
best_models <- purrr::map(tuned_models, ~ select_best(.x))
```

4. finalize and fits the models to the training set
```{r}
final_models <- purrr::map2(workflows_list, best_models, ~ finalize_workflow(.x, .y) |> fit(data = train_data))
```

5. evaluation on the test set
```{r}
results <- purrr::map(final_models, ~ {
  predict(.x, test_data) |>
    bind_cols(test_data) |>
    metrics(truth = spotify_track_popularity, estimate = .pred)})

# summary RMSE
results_df <- bind_rows(results, .id = "model") |>
  filter(.metric == "rmse") |> arrange(.estimate)

print(results_df)
```

```{r}
all_predictions <- purrr::imap(final_models, ~ {
  predict(.x, test_data) |>
    bind_cols(test_data) |>
    mutate(model = .y)
}) |> bind_rows()

```

```{r}
ggplot(all_predictions, aes(x = spotify_track_popularity, y = .pred)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~model, scales = "free") +
  labs(
    title = "Predictions vs Real Values for Models",
    x = "Real Values (spotify_track_popularity)",
    y = "Predictions"
  ) +
  theme_minimal()
```

features importance
```{r}
# VIP plot list
vip_plots <- list()

# Random Forest
rf_fit <- extract_fit_parsnip(final_models$rf)
if (inherits(rf_fit$fit, "randomForest")) {
  vip_plots$rf <- vip(rf_fit$fit, num_features = 15) + ggtitle("Random Forest - Variable Importance")
}

# XGBoost
xgb_fit <- extract_fit_parsnip(final_models$xgb)
if (inherits(xgb_fit$fit, "xgb.Booster")) {
  vip_plots$xgb <- vip(xgb_fit$fit, num_features = 15) + ggtitle("XGBoost - Variable Importance")
}

# visualize plots
library(patchwork)

# show them
if (length(vip_plots) == 2) {
  vip_plots$rf + vip_plots$xgb
} else {
  # if only one is possible
  vip_plots[[1]]
}
```

QQ plot
```{r}
numeric_data <- regression_data |> select(where(is.numeric))
par(mfrow = c(2,2))  # cambia a seconda del numero di variabili

for (colname in names(numeric_data)) {
  qqnorm(numeric_data[[colname]], main = paste("QQ Plot -", colname))
  qqline(numeric_data[[colname]])
}
```

## Sono questi due periodi diversi?? Guardiamo i primi due elementi
```{r}
loudness_summary <- full_data_clean %>%
  group_by(year) %>%
  summarise(
    mean_loudness = mean(loudness, na.rm = TRUE),
    sd_loudness = sd(loudness, na.rm = TRUE),
    n = n(),
    se = sd_loudness / sqrt(n)  # standard error
  )

ggplot(loudness_summary, aes(x = year, y = mean_loudness)) +
  geom_line(color = "steelblue", size = 1) +
  geom_ribbon(aes(ymin = mean_loudness - se, ymax = mean_loudness + se), 
              fill = "lightblue", alpha = 0.4) +
  labs(title = "Loudness Through the Years",
       x = "Year", y = "Average Loudness (dB)") +
  theme_minimal()
```


```{r}



```

