---
title: "R Notebook"
output: html_notebook
---

# Spotify Project

Authors: [Bernacchia Alessia](https://github.com/AlessiaBernacchia), [Pioda Tommaso](https://github.com/Thetommigun432), [Villani Giacomo](https://github.com/DownToTheGround)

```{r}
library(vip)
library(readr)
library(dplyr)
library(tidymodels)
library(ggplot2)
library(tidymodels)
library(rsample)
library(tidyr)
library(stringr)
library(purrr)
library(caret)
library(doParallel)
library(ggcorrplot)
library(finetune)
#install.packages("xgboost")
library(parsnip)
library(randomForest)
library(kknn)
library(xgboost)

# for classification
library(pROC)
```

*Functions*:

```{r}
# count NaN for each column
count_na_per_column <- function(data) {
  sapply(data, function(col) sum(is.na(col)))}

count_na_per_column_df <- function(data) {
  na_counts <- sapply(data, function(col) sum(is.na(col)))
  na_counts_df <- data.frame(
    feature = names(na_counts),
    na_count = na_counts
  )
  na_counts_df %>% arrange(desc(na_count))
}

# extract first genre for the spotify genre
extract_first_genre <- function(df, genre_column = "spotify_genre") {
  # check if column exists
  if (!genre_column %in% colnames(df)) {
    stop(paste("La colonna", genre_column, "non esiste nel dataframe"))
  }
  
  # removes useless not required characters
  df[[genre_column]] <- gsub("\\[|\\]|'", "", df[[genre_column]])
  
  # create a list, with div = ,
  genre_list <- strsplit(df[[genre_column]], split = ",")
  
  # extract only the first genre for each line or rename "others"
  df$first_genre <- sapply(genre_list, function(x) {
    x <- trimws(x)  # remove blanks
    if (length(x) == 0 || is.na(x[1]) || x[1] == "") {
      return("others")
    } else {
      return(x[1])
    }
  })
  
  # convert to categorical value
  df$first_genre <- as.factor(df$first_genre)
  
  # output
  return(df)
}

# process and encode the genre column
process_and_encode_genre <- function(df, genre_column = "first_genre", top_n = 15) {
  # check if column exists
  if (!genre_column %in% colnames(df)) {
    stop(paste("La colonna", genre_column, "non esiste nel dataframe"))
  }
  
  # count the frequency
  genre_counts <- table(df[[genre_column]])
  
  # put them in decrescent order
  genre_counts_sorted <- sort(genre_counts, decreasing = TRUE)
  
  # take top N genres
  top_genres <- names(genre_counts_sorted)[1:min(top_n, length(genre_counts_sorted))]
  
  # rename not in top N into 'others'
  df[[genre_column]] <- ifelse(
    df[[genre_column]] %in% top_genres,
    df[[genre_column]],
    "others"
  )
  
  # convert
  df[[genre_column]] <- as.factor(df[[genre_column]])
  
  # remove original column
  if ("spotify_genre" %in% colnames(df)) {
    df <- subset(df, select = -c(spotify_genre))
  }
  
  # create dummy variables (one-hot encoding)
  dummy <- dummyVars(paste("~", genre_column), data = df)
  dummy_df <- data.frame(predict(dummy, newdata = df))
  
  final_df <- bind_cols(
    df %>% select(-all_of(genre_column)),
    dummy_df
  )
  
  # output
  return(final_df)
}

# encode popularity column
encode_spotify_popularity <- function(df, column = "spotify_track_popularity") {
  # check if column exists
  if (!column %in% colnames(df)) {
    stop(paste("La colonna", column, "non esiste nel dataframe"))
  }
  
  categorize_popularity <- function(x) {
    if (is.na(x)) return(NA)
    if (x <= 25) {
      return("low")
    } else if (x <= 50) {
      return("medium")
    } else if (x <= 75) {
      return("high")
    } else if (x <= 100) {
      return("very high")
    } else {
      return(NA)  # out of range
    }
  }
  
  df[[column]] <- sapply(df[[column]], categorize_popularity)
  
  # convert in ordered factor
  df[[column]] <- factor(df[[column]], 
                         levels = c("low", "medium", "high", "very high"), 
                         ordered = TRUE)
  
  # output
  return(df)
}

# plots the confusion matrix
plot_conf_matrix <- function(conf_matrix_object, title = "Confusion Matrix") {
  
  # converts confusion matrix into dataframe
  cm_df <- as.data.frame(conf_matrix_object$table)
  
  # find columns names (Truth, Prediction, Freq)
  colnames(cm_df) <- tolower(colnames(cm_df)) # rende tutto minuscolo
  
  # rename them
  names(cm_df)[names(cm_df) %in% c("reference", "truth")] <- "Actual"
  names(cm_df)[names(cm_df) == "prediction"] <- "Predicted"
  names(cm_df)[names(cm_df) %in% c("freq", "n")] <- "Freq"
  
  # plot
  p <- ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Freq)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = Freq), size = 5, color = "black") +
    scale_fill_gradient(low = "white", high = "steelblue") +
    labs(title = title,
         x = "Predicted Class",
         y = "Actual Class",
         fill = "Count") +
    theme_minimal(base_size = 14) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  return(p)
}
```

## Import the data

```{r}
billboard <- read_csv("billboard.csv", 
                           col_types = cols(
                            url = col_skip(), 
                            song = col_skip(),
                            instance = col_skip()
                            ))
print(head(billboard,5))
```

```{r}
audio_features <- read_csv("audio_features.csv", 
                          col_types = cols( 
                            performer = col_skip(), 
                            song = col_skip(), 
                            spotify_track_preview_url = col_skip(),
                            spotify_track_explicit = col_skip(),
                            spotify_track_album = col_skip(),
                            spotify_track_id = col_skip()
                            ))
print(head(audio_features,5))
```

```{r}
billboard <- na.omit(billboard) # remove all NaN
audio_features <- as_tibble(audio_features) # translate in a modern and more flexible version
audio_features <- na.omit(audio_features) # remove all NaN

```

```{r}
full_data <- left_join(billboard, audio_features, by = "song_id") # merge datasets
```

## Data exploration and cleaning

```{r}
glimpse(full_data)
```

### Missing Values

```{r}
na_counts <- count_na_per_column_df(full_data)
print(na_counts)
```

```{r}
full_data_clean <- full_data
full_data_clean$previous_week_position[is.na(full_data_clean$previous_week_position)] <- 0 # substitute NaN with 0
full_data_clean <- na.omit(full_data_clean) # remove all NaN

nrow(full_data_clean)
ncol(full_data_clean) 

```

### Distributions

Distribution of numeric data

```{r}
numeric_data <- full_data_clean |> select(where(is.numeric))

numeric_data_long <- numeric_data |> pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# histograms distribution facet
ggplot(numeric_data_long, aes(x = value)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 25) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +  # 3 columns
  theme_minimal()
#ggsave("0_numeric_histogram_distribution.png", width = 12, height = 16)
```

Distribution of categorical data (due the fact there are too many unique values, we choose to filter and observe only the first 50 most frequent variables)

```{r}
categorical_data <- full_data_clean |> select(where(~ !is.numeric(.)))

categorical_data_long <- categorical_data |> pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# select top 50
categorical_data_top50 <- categorical_data_long |>
  group_by(variable, value) |>
  summarise(count = n(), .groups = "drop") |>
  group_by(variable) |>
  slice_max(order_by = count, n = 50)

# barplot distribution facet
ggplot(categorical_data_top50, aes(x = value, y = count)) +
  geom_col(fill = "steelblue", color = "white") + 
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  theme_minimal() +
  theme(#axis.text.x = element_text(angle = 45, hjust = 1))
    axis.text.x = element_blank() )
#ggsave("0_categorical_histogram_distribution.png", width = 12, height = 16)
```

### Cleaning

```{r}
hist(full_data_clean$instrumentalness)
```

```{r}
hist(full_data_clean$time_signature)
```

```{r}
hist(full_data_clean$speechiness)
```

```{r}
hist(full_data_clean$spotify_track_duration_ms)
```

From the distribution of `instrumentalness` and `time signature` they seem to be not significant for our purpose. Then we decide to remove them.

```{r}
full_data_clean <- full_data_clean %>% select(-c(instrumentalness, time_signature))
```

### Outliers

```{r}
# histograms distribution facet
ggplot(numeric_data_long, aes(x = value, y = "")) +
  geom_boxplot(fill = "steelblue") +
  facet_wrap(~ variable, scales = "free", ncol = 3) + 
  theme_minimal()
#ggsave("0_numeric_boxplot_distribution.png", width = 12, height = 16)
```

```{r}
# correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

ggcorrplot(cor_matrix, 
           method = "square",      # o "circle"
           type = "lower",         # lower triangle        
           colors = c("red", "white", "blue"),
           title = "Correlation Matrix",
           ggtheme = ggplot2::theme_minimal())
#ggsave("0_numeric_heatmap.png", width = 12, height = 16)
```

```{r}
# new features creation
billboard_summary <- billboard %>%
  group_by(song_id) %>%
  summarise(
    min_position = min(week_position),
    max_position = max(week_position),
    avg_position = mean(week_position),
    weeks_in_chart = n_distinct(week_id),
    final_position = last(week_position)
  )

full_data_aggregate <- merge(billboard_summary, audio_features, by = "song_id", all = FALSE)
full_data_aggregate <- full_data_aggregate %>% select(-c(instrumentalness, time_signature))

```

```{r}
print(head(full_data_aggregate, 5))
```

```{r}
nrow(billboard)
nrow(billboard_summary)
nrow(audio_features)
nrow(full_data)
nrow(full_data_aggregate)
```

## Question 1 - Given the audio features of a new song, in which position it will chart on Billboard?

### Regression - predict the position on the Billboard

#### Preprocessing

Remove unrelevant features

```{r}
regression_data <- full_data_aggregate %>% select(-c(song_id, min_position, max_position, avg_position, weeks_in_chart, final_position))
print(head(regression_data,5))
```



select only important genres (\>=200 density)

```{r}
# # count frequency of each sample in first_genre column
# genre_counts <- table(regression_data$first_genre)
# 
# # identify genres with less than 200 samples
# low_count_genres <- names(genre_counts[genre_counts < 300])
# 
# # substitute those genres with "others"
# regression_data$first_genre <- ifelse(regression_data$first_genre %in% low_count_genres, "others", regression_data$first_genre)
# 
# # verify the counts after grouping
# genre_counts_after_grouping <- table(regression_data$first_genre)
# 
# # drop original spotify_genre column
# regression_data <- subset(regression_data, select = -c(spotify_genre))
# 
# # ensure the column is upgraded for dummyVars
# df <- data.frame(first_genre = regression_data$first_genre)
# 
# # create dummy variables (one-hot encoding)
# dummy <- dummyVars(~ first_genre, data = df)
# final_df <- data.frame(predict(dummy, newdata = df))
# 
# # bind dummy variables to the rest of regression_data
# regression_data_encoded_1 <- bind_cols(
#   regression_data %>% select(-first_genre),
#   final_df
# )
```

One hot encoding of the 15 most frequent genre

```{r}
regression_data <- extract_first_genre(regression_data)
regression_data_encoded <- process_and_encode_genre(regression_data, top_n = 15)
```

```{r}
# # frequency counter
# genre_counts <- table(regression_data$first_genre)
# 
# # order based in decrescent order the frequency
# genre_counts_sorted <- sort(genre_counts, decreasing = TRUE)
# 
# # take only first 15 genres
# top_15_genres <- names(genre_counts_sorted)[1:15]
# 
# # named others as "others"
# regression_data$first_genre <- ifelse(
#   regression_data$first_genre %in% top_15_genres,
#   regression_data$first_genre,
#   "others"
# )
# 
# # verify
# genre_counts_after_grouping <- table(regression_data$first_genre)
# 
# # drop original spotify_genre column
# regression_data <- subset(regression_data, select = -c(spotify_genre))
# 
# # ensure the column is upgraded for dummyVars
# df <- data.frame(first_genre = regression_data$first_genre)
# 
# # create dummy variables (one-hot encoding)
# dummy <- dummyVars(~ first_genre, data = df)
# final_df <- data.frame(predict(dummy, newdata = df))
# 
# # bind dummy variables to the rest of regression_data (without first_genre)
# regression_data_encoded <- bind_cols(
#   regression_data %>% select(-first_genre),
#   final_df
# )


print(head(regression_data_encoded, 5))
```

#### Regression

```{r}

set.seed(45)
split <- initial_split(head(regression_data_encoded,100), prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)

# recipe
#recipe_spec <- recipe(spotify_track_popularity ~ ., data = train_data)
recipe_spec <- recipe(spotify_track_popularity ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

# define models
task = 'regression'

models <- list(
  rf = rand_forest(mtry = 5) %>% set_engine("randomForest") %>% set_mode(task),
  #lm = linear_reg() %>% set_engine("lm") %>% set_mode(task),
  xgb = boost_tree(trees = 1000, tree_depth = 10) %>% set_engine("xgboost") %>% set_mode(task),
  knn = nearest_neighbor(neighbors = 5) %>% set_engine("kknn") %>% set_mode(task)
)

# workflow list
workflows_list <- purrr::map(models, ~ workflow() %>% add_model(.x) %>% add_recipe(recipe_spec))

# train the models
fitted_models <- purrr::map(workflows_list, ~ fit(.x, data = train_data))

# evaluation
results <- purrr::map(fitted_models, ~ {
  predict(.x, test_data) %>%
    bind_cols(test_data) %>%
    metrics(truth = spotify_track_popularity, estimate = .pred)
})

# results
results_df <- bind_rows(results, .id = "model")

results_df %>%
  filter(.metric == "rmse") %>%
  arrange(.estimate)

print(results_df)
```

```{r}
results <- list()

for (model_name in names(fitted_models)) {
  model <- models[[model_name]]
  
  wf <- workflow() %>%
    add_model(model) %>%
    add_recipe(recipe_spec)
  
  fit_model <- fit(wf, data = train_data)
  
  preds <- predict(fit_model, test_data) %>%
    bind_cols(test_data) %>%
    mutate(model = model_name)
  
  results[[model_name]] <- preds
}

all_predictions <- bind_rows(results)

```

```{r}

ggplot(all_predictions, aes(x = spotify_track_popularity, y = .pred)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~model, scales = "free") +
  labs(
    title = "Predictions vs Real Values",
    x = "Real Values (spotify_track_popularity)",
    y = "Predictions"
  ) +
  theme_minimal()
```

```{r}
# VIP plot list
vip_plots <- list()

# Random Forest
rf_fit <- extract_fit_parsnip(fitted_models$rf)
if (inherits(rf_fit$fit, "randomForest")) {
  vip_plots$rf <- vip(rf_fit$fit, num_features = 15) + ggtitle("Random Forest - Variable Importance")
}

# XGBoost
xgb_fit <- extract_fit_parsnip(fitted_models$xgb)
if (inherits(xgb_fit$fit, "xgb.Booster")) {
  vip_plots$xgb <- vip(xgb_fit$fit, num_features = 15) + ggtitle("XGBoost - Variable Importance")
}

# visualize plots
library(patchwork)

# show them
if (length(vip_plots) == 2) {
  vip_plots$rf + vip_plots$xgb
} else {
  # if only one is possible
  vip_plots[[1]]
}

```

### Classification- predict which level of popularity

#### Preprocessing

```{r}
classification_data <- full_data_aggregate %>% select(-c( song_id, min_position, max_position, avg_position, weeks_in_chart, final_position))
print(head(classification_data,5))
```
Encode popularity
```{r}
classification_data <- encode_spotify_popularity(classification_data)

classification_data <- classification_data |>
  filter(!is.na(spotify_track_popularity)) 
```
Encode genre
```{r}
classification_data <- extract_first_genre(classification_data)
classification_data <- process_and_encode_genre(classification_data, top_n = 15)
```

```{r}
#print number of NaN for each column
count_na_per_column_df(classification_data)
```
```{r}
print(head(classification_data,5))
```

```{r}
# select only numerical data
numeric_data_clss <- classification_data %>%
  select(where(is.numeric))

# correlation matrix
cor_matrix <- cor(numeric_data_clss, use = "complete.obs")

ggcorrplot(cor_matrix, 
           method = "square",     # or "circle"
           type = "lower",        # only lower triangle        
           colors = c("red", "white", "blue"),  # come in seaborn
           title = "Correlation Matrix",
           ggtheme = ggplot2::theme_minimal())
```

```{r}
ggplot(classification_data, aes(x = spotify_track_popularity)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Popularity Distribution",
       x = "Spotify Track Popularity",
       y = "Counter") +
  theme_minimal()
```

```{r}
set.seed(42)
spotify_train_test <- classification_data |> initial_split(prop =  0.7)
spotify_train <- training(spotify_train_test) 
spotify_test  <- testing(spotify_train_test)
```
#### Classification
Random forest
```{r}
rand_forest_spec <- 
  rand_forest(mtry = 5, trees = 200) |>  
  set_engine('randomForest') |> 
  set_mode('classification')

rf_recipe = recipe(spotify_track_popularity~.,
                   data=spotify_train)

rf_wf <- workflow() |> 
  add_recipe(rf_recipe) |> 
  add_model(rand_forest_spec)

rf_wf_fit <- rf_wf |> 
  fit(spotify_train)

rf_wf_fit
```

```{r}
spotify_test_rf <- rf_wf_fit |> 
  augment(spotify_test)

# confusion matrix
cm_rF <- spotify_test_rf |> 
  conf_mat(truth = spotify_track_popularity,
           estimate = .pred_class)

# visualize
plot_conf_matrix(cm_rF, title = "Confusion Matrix - Classification Random Forest")

```


```{r}
head(summary(cm_rF),4)
```

```{r}
# select(spotify_test_rf, starts_with(".pred_"))

rf_wf_fit |> 
  extract_fit_engine() |> 
  vip()
```
## Question 2 - As a manager, should I continue to invest in this song given its statistics? Will it remain popular over time?

### Regression - predict the position on the Billboard

#### Preprocessing

Remove unrelevant features

```{r}
regression_data <- full_data_aggregate %>% select(-c(song_id))
print(head(regression_data,5))
```
one hot encoding of the most 15 important genres
```{r}
regression_data <- extract_first_genre(regression_data)
regression_data_encoded <- process_and_encode_genre(regression_data, top_n = 15)
```

```{r}
print(head(regression_data_encoded, 5))
```

#### Regression

```{r}
set.seed(45)
split <- initial_split(head(regression_data_encoded,100), prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)

# recipe
recipe_spec <- recipe(final_position ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

task = 'regression'

# models
models <- list(
  rf = rand_forest(mtry = 5) %>% set_engine("randomForest") %>% set_mode(task),
  #lm = linear_reg() %>% set_engine("lm") %>% set_mode(task),
  xgb = boost_tree(trees = 100, tree_depth = 10) %>% set_engine("xgboost") %>% set_mode(task),
  knn = nearest_neighbor(neighbors = 5) %>% set_engine("kknn") %>% set_mode(task)
)

# workflow
workflows_list <- purrr::map(models, ~ workflow() %>% add_model(.x) %>% add_recipe(recipe_spec))

# train models
fitted_models <- purrr::map(workflows_list, ~ fit(.x, data = train_data))

# evaluation
results <- purrr::map(fitted_models, ~ {
  predict(.x, test_data) %>%
    bind_cols(test_data) %>%
    metrics(truth = final_position, estimate = .pred)
})

# results
results_df <- bind_rows(results, .id = "model")

results_df %>%
  filter(.metric == "rmse") %>%
  arrange(.estimate)

print(results_df)
```

```{r}
results <- list()

for (model_name in names(fitted_models)) {
  model <- models[[model_name]]
  
  wf <- workflow() %>%
    add_model(model) %>%
    add_recipe(recipe_spec)
  
  fit_model <- fit(wf, data = train_data)
  
  preds <- predict(fit_model, test_data) %>%
    bind_cols(test_data) %>%
    mutate(model = model_name)
  
  results[[model_name]] <- preds
}

all_predictions <- bind_rows(results)

```

```{r}
ggplot(all_predictions, aes(x = final_position, y = .pred)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~model, scales = "free") +
  labs(
    title = "Predictions vs Real Values per Modello",
    x = "Real Values (spotify_track_popularity)",
    y = "Predictions"
  ) +
  theme_minimal()
```

```{r}
# VIP plot list
vip_plots <- list()

# Random Forest
rf_fit <- extract_fit_parsnip(fitted_models$rf)
if (inherits(rf_fit$fit, "randomForest")) {
  vip_plots$rf <- vip(rf_fit$fit, num_features = 15) + ggtitle("Random Forest - Variable Importance")
}

# XGBoost
xgb_fit <- extract_fit_parsnip(fitted_models$xgb)
if (inherits(xgb_fit$fit, "xgb.Booster")) {
  vip_plots$xgb <- vip(xgb_fit$fit, num_features = 15) + ggtitle("XGBoost - Variable Importance")
}

# visualize plots
library(patchwork)

# show them
if (length(vip_plots) == 2) {
  vip_plots$rf + vip_plots$xgb
} else {
  # if only one is possible
  vip_plots[[1]]
}
```

### Classification- predict whether a sign will be in the top 30

#### Preprocessing

```{r}
classification_data <- full_data_aggregate |> select(-c( song_id))
print(head(classification_data,5))
```
Encode popularity
```{r}
classification_data <- encode_spotify_popularity(classification_data)

classification_data <- classification_data |>
  filter(!is.na(spotify_track_popularity)) 
```
Encode genre
```{r}
classification_data <- extract_first_genre(classification_data)
classification_data <- process_and_encode_genre(classification_data, top_n = 5)
```

```{r}
#print number of NaN for each column
count_na_per_column_df(classification_data)
```




```{r}
#create the new variable top30: whether a song is in the top 30
classification_data <- classification_data |>
  mutate(top30 = ifelse(final_position <= 30, 1, 0))
# remove final_position
classification_data <- classification_data |> select(-c( final_position))
print(head(classification_data,5))
```

```{r}
# select only numerical data
numeric_data_clss <- classification_data |>
  select(where(is.numeric))

# correlation matrix
cor_matrix <- cor(numeric_data_clss, use = "complete.obs")

ggcorrplot(cor_matrix, 
           method = "square",     # or "circle"
           type = "lower",        # only lower triangle        
           colors = c("red", "white", "blue"),
           title = "Correlation Matrix",
           ggtheme = ggplot2::theme_minimal())
```

```{r}
hist(classification_data$top30)
```
#### Classification
```{r}
set.seed(42)
classification_data$top30 <- as.factor(classification_data$top30)
spotify_train_test <- classification_data |> initial_split(prop =  0.7)
spotify_train <- training(spotify_train_test) 
spotify_test  <- testing(spotify_train_test)
```

```{r}
rand_forest_spec <- 
  rand_forest(mtry = 5, trees = 200) |>  
  set_engine('randomForest') |> 
  set_mode('classification')

rf_recipe = recipe(top30~.,
                   data=spotify_train)

rf_wf <- workflow() |> 
  add_recipe(rf_recipe) |> 
  add_model(rand_forest_spec)

rf_wf_fit <- rf_wf |> 
  fit(spotify_train)

rf_wf_fit
```

```{r}
spotify_test_rf <- rf_wf_fit |> 
  augment(spotify_test)

# confusion matrix
cm_rF <- spotify_test_rf |> 
  conf_mat(truth = top30,
           estimate = .pred_class)

# visualize
plot_conf_matrix(cm_rF, title = "Confusion Matrix - Classification Random Forest")
```

```{r}
head(summary(cm_rF),4)
```

```{r}
roc_rF<-roc_curve(spotify_test_rf,
          top30,
          .pred_0,event_level = "first")
roc_rF |> autoplot()
```

```{r}
rf_wf_fit |> 
  extract_fit_engine() |> 
  vip()
```

```{r}
df_filtered <- full_data_clean |>
  select(colnames(full_data_clean),song_id,week_id)

df_filtered$year <- as.numeric(sub(".*/.*/", "", df_filtered$week_id))

df_filtered_2000_15 <- df_filtered[df_filtered$year >= 2015 & df_filtered$year < 2025, ]

df_filtered_2000_15$week_id <- NULL

df_filtered_2000_15 <- unique(df_filtered_2000_15)

df_filtered_2000_15_genre <- extract_first_genre(df_filtered_2000_15)
df_filtered_2000_15_genre_encoded <- process_and_encode_genre(df_filtered_2000_15_genre, top_n = 25)

df_filtered_2000_15_genre_encoded$spotify_track_duration_min <- df_filtered_2000_15_genre_encoded$spotify_track_duration_ms/60000

df_filtered_2000_15_genre_encoded$spotify_track_duration_ms <- NULL
```

```{r}
df_filtered_2000_15_genre_encoded$week_position <- NULL
df_filtered_2000_15_genre_encoded$performer <- NULL
df_filtered_2000_15_genre_encoded$song_id <- NULL
df_filtered_2000_15_genre_encoded$previous_week_position <- NULL
df_filtered_2000_15_genre_encoded$peak_position <- NULL
df_filtered_2000_15_genre_encoded$weeks_on_chart <- NULL
df_filtered_2000_15_genre_encoded$year <- NULL
```

```{r}
df_filtered_2000_15_genre_encoded <- unique(df_filtered_2000_15_genre_encoded)
df_filtered_2000_15_genre_encoded
```



```{r}
registerDoParallel(cores = parallel::detectCores() - 1)

set.seed(42)

# 1. Split
split <- initial_split(df_filtered_2000_15_genre_encoded, prop = 0.8)
#split <- initial_split(regression_data_encoded_scaled, prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)

# 2. Recipe
recipe_spec <- recipe(spotify_track_popularity ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

# 3. Resampling
cv_folds <- vfold_cv(train_data, v = 5)

# 4. Definisci i modelli con parametri da tunare
models <- list(
  rf = rand_forest(mtry = tune(), trees = 1000) %>%
    set_engine("randomForest") %>% set_mode("regression"),

  xgb = boost_tree(trees = 1000, tree_depth = tune(), learn_rate = tune()) %>%
    set_engine("xgboost") %>% set_mode("regression"),

  knn = nearest_neighbor(neighbors = tune()) %>%
    set_engine("kknn") %>% set_mode("regression")
)

# 5. Crea i workflow
workflows_list <- purrr::imap(models, function(model, name) {
  workflow() %>%
    add_model(model) %>%
    add_recipe(recipe_spec)
})

# 6. Tuning veloce con tune_race_anova
tuned_models <- purrr::imap(workflows_list, function(wf, name) {
  tune_race_anova(
    wf,
    resamples = cv_folds,
    grid = 5,
    metrics = metric_set(rmse),
    control = control_race(verbose_elim = TRUE)
  )
})

# 7. Estrai i migliori iperparametri
best_models <- purrr::map(tuned_models, ~ select_best(.x))

# 8. Finalizza e fitta
final_models <- purrr::map2(workflows_list, best_models, ~ finalize_workflow(.x, .y) %>% fit(data = train_data))

# 9. Valuta su test set
results <- purrr::map(final_models, ~ {
  predict(.x, test_data) %>%
    bind_cols(test_data) %>%
    metrics(truth = spotify_track_popularity, estimate = .pred)
})

# 10. Riassunto risultati RMSE
results_df <- bind_rows(results, .id = "model") %>%
  filter(.metric == "rmse") %>%
  arrange(.estimate)

print(results_df)
```

```{r}
all_predictions <- purrr::imap(final_models, ~ {
  predict(.x, test_data) %>%
    bind_cols(test_data) %>%
    mutate(model = .y)
}) %>% bind_rows()
```


```{r}
ggplot(all_predictions, aes(x = spotify_track_popularity, y = .pred)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~model, scales = "free") +
  labs(
    title = "Predictions vs model values",
    x = "Real values (spotify_track_popularity)",
    y = "Predictions"
  ) +
  theme_minimal()
```




