---
title: "R Notebook"
output: html_notebook
---

# Spotify Project

Authors: [Bernacchia Alessia](https://github.com/AlessiaBernacchia), [Pioda Tommaso](https://github.com/Thetommigun432), [Villani Giacomo](https://github.com/DownToTheGround)

```{r}
library(vip)
library(readr)
library(dplyr)
library(tidymodels)
library(ggplot2)
library(tidymodels)
library(rsample)
library(tidyr)
library(stringr)
library(purrr)
library(caret)
library(kknn)
library(ggcorrplot)
#install.packages("xgboost")
library(parsnip)
library(randomForest)
library(kknn)
library(xgboost)
```

*Functions*:

```{r}
# count NaN for each column
count_na_per_column <- function(data) {
  sapply(data, function(col) sum(is.na(col)))}

count_na_per_column_df <- function(data) {
  na_counts <- sapply(data, function(col) sum(is.na(col)))
  na_counts_df <- data.frame(
    feature = names(na_counts),
    na_count = na_counts
  )
  na_counts_df %>% arrange(desc(na_count))
}

```

## Import the data

```{r}
billboard <- read_csv("billboard.csv", 
                           col_types = cols(
                            url = col_skip(), 
                            song = col_skip(),
                            instance = col_skip()
                            ))
print(head(billboard,5))
```

```{r}
audio_features <- read_csv("audio_features.csv", 
                          col_types = cols( 
                            performer = col_skip(), 
                            song = col_skip(), 
                            spotify_track_preview_url = col_skip(),
                            spotify_track_explicit = col_skip(),
                            spotify_track_album = col_skip(),
                            spotify_track_id = col_skip()
                            ))
print(head(audio_features,5))
```

```{r}
billboard <- na.omit(billboard) # remove all NaN
audio_features <- as_tibble(audio_features) # translate in a modern and more flexible version
audio_features <- na.omit(audio_features) # remove all NaN

```

```{r}
full_data <- left_join(billboard, audio_features, by = "song_id") # merge datasets
```

## Data exploration and cleaning

```{r}
glimpse(full_data)
```

### Missing Values

```{r}
na_counts <- count_na_per_column_df(full_data)
print(na_counts)
```

```{r}
full_data_clean <- full_data
full_data_clean$previous_week_position[is.na(full_data_clean$previous_week_position)] <- 0 # substitute NaN with 0
full_data_clean <- na.omit(full_data_clean) # remove all NaN

nrow(full_data_clean)
ncol(full_data_clean) 

```

### Distributions

Distribution of numeric data

```{r}
numeric_data <- full_data_clean %>% select(where(is.numeric))

numeric_data_long <- numeric_data %>% pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# histograms distribution facet
ggplot(numeric_data_long, aes(x = value)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 25) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +  # 3 columns
  theme_minimal()
#ggsave("0_numeric_histogram_distribution.png", width = 12, height = 16)
```

Distribution of categorical data (due the fact there are too many unique values, we choose to filter and observe only the first 50 most frequent variables)

```{r}
categorical_data <- full_data_clean %>% select(where(~ !is.numeric(.)))

categorical_data_long <- categorical_data %>% pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# select top 50
categorical_data_top50 <- categorical_data_long %>%
  group_by(variable, value) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(variable) %>%
  slice_max(order_by = count, n = 50)

# barplot distribution facet
ggplot(categorical_data_top50, aes(x = value, y = count)) +
  geom_col(fill = "steelblue", color = "white") + 
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  theme_minimal() +
  theme(#axis.text.x = element_text(angle = 45, hjust = 1))
    axis.text.x = element_blank() )
#ggsave("0_categorical_histogram_distribution.png", width = 12, height = 16)
```

### Cleaning

```{r}
hist(full_data_clean$instrumentalness)
```

```{r}
hist(full_data_clean$time_signature)
```

```{r}
hist(full_data_clean$speechiness)
```

```{r}
hist(full_data_clean$spotify_track_duration_ms)
```

From the distribution of `instrumentalness` and `time signature` they seem to be not significant for our purpose. Then we decide to remove them.

```{r}
full_data_clean <- full_data_clean %>% select(-c(instrumentalness, time_signature))
```

### Outliers

```{r}
# histograms distribution facet
ggplot(numeric_data_long, aes(x = value, y = "")) +
  geom_boxplot(fill = "steelblue") +
  facet_wrap(~ variable, scales = "free", ncol = 3) + 
  theme_minimal()
#ggsave("0_numeric_boxplot_distribution.png", width = 12, height = 16)
```

```{r}
# correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

ggcorrplot(cor_matrix, 
           method = "square",      # o "circle"
           type = "lower",         # solo triangolo inferiore         
           colors = c("red", "white", "blue"),  # come in seaborn
           title = "Matrice di Correlazione",
           ggtheme = ggplot2::theme_minimal())
```

```{r}
# new features creation
billboard_summary <- billboard %>%
  group_by(song_id) %>%
  summarise(
    min_position = min(week_position),
    max_position = max(week_position),
    avg_position = mean(week_position),
    weeks_in_chart = n_distinct(week_id),
    final_position = last(week_position)
  )

full_data_aggregate <- merge(billboard_summary, audio_features, by = "song_id", all = FALSE)
full_data_aggregate <- full_data_aggregate %>% select(-c(instrumentalness, time_signature))

```

```{r}
print(head(full_data_aggregate, 5))
```

```{r}
nrow(billboard)
nrow(billboard_summary)
nrow(audio_features)
nrow(full_data)
nrow(full_data_aggregate)
```

## Regression - predict the position on the billboard

### Preprocessing

Remove unrelevant features

```{r}
regression_data <- full_data_aggregate %>% select(-c(song_id, min_position, max_position, avg_position, weeks_in_chart, final_position))
print(head(regression_data,5))
```

```{r}
# removes useless not required characters
regression_data$spotify_genre <- gsub("\\[|\\]|'", "", regression_data$spotify_genre)
# create a list, with div = ,
regression_data$spotify_genre <- strsplit(regression_data$spotify_genre, split = ",")
# extract only the first genre for each line
regression_data$first_genre <- sapply(regression_data$spotify_genre, function(x) x[1])

regression_data$first_genre <- sapply(regression_data$spotify_genre, function(x) {
  if (length(x) == 0) {
    return("others")
  } else {
    return(x[1])
  }
})

# audio_features$spotify_track_popularity <- sapply(audio_features$spotify_track_popularity, function(x) {
#   if (x <= 25) {
#     return("low")
#   }
#   if (x > 25 & x <= 50) {
#     return("medium")
#   }
#   if (x > 50 & x <= 75) {
#     return("high")
#   }
#     if (x > 75 & x <= 100) {
#     return("very high")
#   } })
```

select only important genres (\>=200 density)

```{r}

# # count frequency of each sample in first_genre column
# genre_counts <- table(regression_data$first_genre)
# 
# # identify genres with less than 200 samples
# low_count_genres <- names(genre_counts[genre_counts < 300])
# 
# # substitute those genres with "others"
# regression_data$first_genre <- ifelse(regression_data$first_genre %in% low_count_genres, "others", regression_data$first_genre)
# 
# # verify the counts after grouping
# genre_counts_after_grouping <- table(regression_data$first_genre)
# 
# # drop original spotify_genre column
# regression_data <- subset(regression_data, select = -c(spotify_genre))
# 
# # ensure the column is upgraded for dummyVars
# df <- data.frame(first_genre = regression_data$first_genre)
# 
# # create dummy variables (one-hot encoding)
# dummy <- dummyVars(~ first_genre, data = df)
# final_df <- data.frame(predict(dummy, newdata = df))
# 
# # bind dummy variables to the rest of regression_data
# regression_data_encoded_1 <- bind_cols(
#   regression_data %>% select(-first_genre),
#   final_df
# )
```

```{r}
# frequency counter
genre_counts <- table(regression_data$first_genre)

# order based in descrescent order the frequency
genre_counts_sorted <- sort(genre_counts, decreasing = TRUE)

# take only first 15 genres
top_15_genres <- names(genre_counts_sorted)[1:15]

# named others as "others"
regression_data$first_genre <- ifelse(
  regression_data$first_genre %in% top_15_genres,
  regression_data$first_genre,
  "others"
)
```

```{r}
print(head(regression_data,5))
```

```{r}

# verify
genre_counts_after_grouping <- table(regression_data$first_genre)

# drop original spotify_genre column
regression_data <- subset(regression_data, select = -c(spotify_genre))

# ensure the column is upgraded for dummyVars
df <- data.frame(first_genre = regression_data$first_genre)

# create dummy variables (one-hot encoding)
dummy <- dummyVars(~ first_genre, data = df)
final_df <- data.frame(predict(dummy, newdata = df))

# bind dummy variables to the rest of regression_data (without first_genre)
regression_data_encoded <- bind_cols(
  regression_data %>% select(-first_genre),
  final_df
)

```

```{r}
print(head(regression_data_encoded, 5))
```

### Regression

```{r}

set.seed(45)
split <- initial_split(regression_data_encoded, prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)

# 1. Prepara il recipe (già fatto)
#recipe_spec <- recipe(spotify_track_popularity ~ ., data = train_data)
recipe_spec <- recipe(spotify_track_popularity ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

task = 'regression'

# 2. Definisci i modelli
models <- list(
  rf = rand_forest(mtry = 5) %>% set_engine("randomForest") %>% set_mode(task),
  #lm = linear_reg() %>% set_engine("lm") %>% set_mode(task),
  xgb = boost_tree(trees = 1000, tree_depth = 10) %>% set_engine("xgboost") %>% set_mode(task),
  knn = nearest_neighbor(neighbors = 5) %>% set_engine("kknn") %>% set_mode(task)
)

# 3. Crea una lista di workflow
workflows_list <- purrr::map(models, ~ workflow() %>% add_model(.x) %>% add_recipe(recipe_spec))

# 4. Fitta tutti i modelli
fitted_models <- purrr::map(workflows_list, ~ fit(.x, data = train_data))

# 5. Valuta su test set
results <- purrr::map(fitted_models, ~ {
  predict(.x, test_data) %>%
    bind_cols(test_data) %>%
    metrics(truth = spotify_track_popularity, estimate = .pred)
})

# 6. Combina risultati in una tabella leggibile
results_df <- bind_rows(results, .id = "model")

results_df %>%
  filter(.metric == "rmse") %>%
  arrange(.estimate)

print(results_df)
```

```{r}
results <- list()

for (model_name in names(fitted_models)) {
  model <- models[[model_name]]
  
  wf <- workflow() %>%
    add_model(model) %>%
    add_recipe(recipe_spec)
  
  fit_model <- fit(wf, data = train_data)
  
  preds <- predict(fit_model, test_data) %>%
    bind_cols(test_data) %>%
    mutate(model = model_name)
  
  results[[model_name]] <- preds
}

# Unione di tutte le predizioni
all_predictions <- bind_rows(results)

```

```{r}

ggplot(all_predictions, aes(x = spotify_track_popularity, y = .pred)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~model, scales = "free") +
  labs(
    title = "Predizioni vs Valori Reali per Modello",
    x = "Valori Reali (spotify_track_popularity)",
    y = "Predizioni"
  ) +
  theme_minimal()
```

```{r}
#vip(fitted_models)
```

```{r}
# if (inherits(fitted_models$rf$fit, "randomForest")) {
#   vip(fitted_models$rf$fit)
# }
```

## Classification- predict whether a sogn will be in the top 10

### Preprocessing

```{r}
classification_data <- full_data_aggregate %>% select(-c( song_id, final_position))
print(head(classification_data,5))
```

```{r}
#create the new variable top10: whether a song is in the top 30
classification_data <- classification_data %>%
  filter(!is.na(spotify_track_popularity)) 

classification_data <- classification_data %>%
  mutate(top30 = ifelse(avg_position <= 30, 1, 0))

# encode popularity
classification_data$spotify_track_popularity <- sapply(classification_data$spotify_track_popularity, function(x) {
  if (x <= 25) {
    return("low")
  }
  if (x > 25 & x <= 50) {
    return("medium")
  }
  if (x > 50 & x <= 75) {
    return("high")
  }
    if (x > 75 & x <= 100) {
    return("very high")
  } 
})
```

```{r}
#remove useless features (usefull only)
classification_data <- classification_data %>% select(-c( avg_position, min_position, max_position))
```

```{r}
# Step 1: Rimuovi caratteri indesiderati
classification_data$spotify_genre <- gsub("\\[|\\]|'", "", classification_data$spotify_genre)

# Step 2: Creare una lista di generi separati da virgola
classification_data$spotify_genre <- strsplit(classification_data$spotify_genre, split = ",")

# Step 3: Estrarre solo il primo genere per ogni riga
classification_data$first_genre <- sapply(classification_data$spotify_genre, function(x) x[1])

classification_data$first_genre <- sapply(classification_data$spotify_genre, function(x) {
  if (length(x) == 0) {
    return("others")
    } else {
    return(x[1])
  }
})

# frequency counter
genre_counts <- table(classification_data$first_genre)

# order based in descrescent order the frequency
genre_counts_sorted <- sort(genre_counts, decreasing = TRUE)

# take only first 15 genres
top_15_genres <- names(genre_counts_sorted)[1:15]

# named others as "others"
classification_data$first_genre <- ifelse(
  classification_data$first_genre %in% top_15_genres,
  classification_data$first_genre,
  "others"
)

# verify
genre_counts_after_grouping <- table(classification_data$first_genre)

# drop original spotify_genre column
classification_data <- subset(classification_data, select = -c(spotify_genre))

# ensure the column is upgraded for dummyVars
df <- data.frame(first_genre = classification_data$first_genre)

# create dummy variables (one-hot encoding)
dummy <- dummyVars(~ first_genre, data = df)
final_df <- data.frame(predict(dummy, newdata = df))

# bind dummy variables to the rest of regression_data (without first_genre)
classification_data <- bind_cols(
  classification_data %>% select(-first_genre),
  final_df
)
```

```{r}
#print number of NaN for each column
count_na_per_column_df(classification_data)
```

```{r}
# select only numerical data
numeric_data_clss <- classification_data %>%
  select(where(is.numeric))

# correlation matrix
cor_matrix <- cor(numeric_data_clss, use = "complete.obs")

ggcorrplot(cor_matrix, 
           method = "square",     # or "circle"
           type = "lower",        # only lower triangle        
           colors = c("red", "white", "blue"),  # come in seaborn
           title = "Correlation Matrix",
           ggtheme = ggplot2::theme_minimal())
```

```{r}
hist(classification_data$top30)
```

```{r}
set.seed(42)
classification_data$top30 <- as.factor(classification_data$top30)
spotify_train_test <- classification_data |> initial_split(prop =  0.7)
spotify_train <- training(spotify_train_test); 
spotify_test  <- testing(spotify_train_test)
```

```{r}
rand_forest_spec <- 
  rand_forest(mtry = 5, trees = 200) |>  
  set_engine('randomForest') |> 
  set_mode('classification')

rf_recipe = recipe(top30~.,
                   data=spotify_train)

rf_wf <- workflow() |> 
  add_recipe(rf_recipe) |> 
  add_model(rand_forest_spec)

rf_wf_fit <- rf_wf |> 
  fit(spotify_train)

rf_wf_fit
```

```{r}
spotify_test_rf <- rf_wf_fit |> 
  augment(spotify_test)

cm_rF <- spotify_test_rf |> 
  conf_mat(truth=top30,
           estimate=.pred_class)
cm_rF
```

```{r}
head(summary(cm_rF),4)
```

```{r}
roc_rF<-roc_curve(spotify_test_rf,
          top30,
          .pred_0,event_level = "first")
roc_rF |> autoplot()
```

```{r}
rf_wf_fit |> 
  extract_fit_engine() |> 
  vip()
```
