---
title: "R Notebook"
output: html_notebook
---

# Spotify Project

In this project we will analyse the evolution of the music, from 1956 until now.

-   First we start to explore whether is possible to predict the popularity of a song from its audio features

-   In second

link to the **dataset**: [jonhthgeek-DataSetSpotify](https://github.com/rfordatascience/tidytuesday/blob/main/data/2021/2021-09-14/readme.md)

**Authors**: [Bernacchia Alessia](https://github.com/AlessiaBernacchia), [Pioda Tommaso](https://github.com/Thetommigun432), [Villani Giacomo](https://github.com/DownToTheGround)

```{r}
# samples for regression
n_samples_to_consider = 100
```


```{r}
library(vip)
library(readr)
library(dplyr)
library(tidymodels)
library(ggplot2)
library(rsample)
library(tidyr)
library(themis)
library(stringr)
library(purrr)
library(caret)
library(kknn)

#install.packages("xgboost")
library(parsnip)
library(randomForest)
library(kknn)
library(xgboost)
library(finetune)
library(doParallel)
library(robustHD)
library(parsnip)

# for exploration
library(corrplot)
library(ggcorrplot)

# for classification
library(pROC)
```

*Functions*:

```{r}
# count NaN for each column
count_na_per_column <- function(data) {
  sapply(data, function(col) sum(is.na(col)))}

count_na_per_column_df <- function(data) {
  na_counts <- sapply(data, function(col) sum(is.na(col)))
  na_counts_df <- data.frame(
    feature = names(na_counts),
    na_count = na_counts
  )
  na_counts_df %>% arrange(desc(na_count))
}

# extract first genre for the spotify genre
extract_first_genre <- function(df, genre_column = "spotify_genre") {
  # check if column exists
  if (!genre_column %in% colnames(df)) {
    stop(paste("La colonna", genre_column, "non esiste nel dataframe"))
  }
  
  # removes useless not required characters
  df[[genre_column]] <- gsub("\\[|\\]|'", "", df[[genre_column]])
  
  # create a list, with div = ,
  genre_list <- strsplit(df[[genre_column]], split = ",")
  
  # extract only the first genre for each line or rename "others"
  df$first_genre <- sapply(genre_list, function(x) {
    x <- trimws(x)  # remove blanks
    if (length(x) == 0 || is.na(x[1]) || x[1] == "") {
      return("others")
    } else {
      return(x[1])
    }
  })
  
  # convert to categorical value
  df$first_genre <- as.factor(df$first_genre)
  
  # output
  return(df)
}

# process and encode the genre column
process_and_encode_genre <- function(df, genre_column = "first_genre", top_n = 15) {
  # check if column exists
  if (!genre_column %in% colnames(df)) {
    stop(paste("La colonna", genre_column, "non esiste nel dataframe"))
  }
  
  # count the frequency
  genre_counts <- table(df[[genre_column]])
  
  # put them in decrescent order
  genre_counts_sorted <- sort(genre_counts, decreasing = TRUE)
  
  # take top N genres
  top_genres <- names(genre_counts_sorted)[1:min(top_n, length(genre_counts_sorted))]
  
  # rename not in top N into 'others'
  df[[genre_column]] <- ifelse(
    df[[genre_column]] %in% top_genres,
    df[[genre_column]],
    "others"
  )
  
  # convert
  df[[genre_column]] <- as.factor(df[[genre_column]])
  
  # remove original column
  if ("spotify_genre" %in% colnames(df)) {
    df <- subset(df, select = -c(spotify_genre))
  }
  
  # create dummy variables (one-hot encoding)
  dummy <- dummyVars(paste("~", genre_column), data = df)
  dummy_df <- data.frame(predict(dummy, newdata = df))
  
  final_df <- bind_cols(
    df %>% select(-all_of(genre_column)),
    dummy_df
  )
  
  # output
  return(final_df)
}

# encode popularity column
encode_spotify_popularity <- function(df, column = "spotify_track_popularity") {
  # check if column exists
  if (!column %in% colnames(df)) {
    stop(paste("La colonna", column, "non esiste nel dataframe"))
  }
  
  categorize_popularity <- function(x) {
    if (is.na(x)) return(NA)
    if (x <= 25) {
      return("low")
    } else if (x <= 50) {
      return("medium")
    } else if (x <= 75) {
      return("high")
    } else if (x <= 100) {
      return("very high")
    } else {
      return(NA)  # out of range
    }
  }
  
  df[[column]] <- sapply(df[[column]], categorize_popularity)
  
  # convert in ordered factor
  df[[column]] <- factor(df[[column]], 
                         levels = c("low", "medium", "high", "very high"), 
                         ordered = TRUE)
  
  # output
  return(df)
}

# plots the confusion matrix
plot_conf_matrix <- function(conf_matrix_object, title = "Confusion Matrix") {
  
  # converts confusion matrix into dataframe
  cm_df <- as.data.frame(conf_matrix_object$table)
  
  # find columns names (Truth, Prediction, Freq)
  colnames(cm_df) <- tolower(colnames(cm_df)) # rende tutto minuscolo
  
  # rename them
  names(cm_df)[names(cm_df) %in% c("reference", "truth")] <- "Actual"
  names(cm_df)[names(cm_df) == "prediction"] <- "Predicted"
  names(cm_df)[names(cm_df) %in% c("freq", "n")] <- "Freq"
  
  # plot
  p <- ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Freq)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = Freq), size = 5, color = "black") +
    scale_fill_gradient(low = "white", high = "steelblue") +
    labs(title = title,
         x = "Predicted Class",
         y = "Actual Class",
         fill = "Count") +
    theme_minimal(base_size = 14) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  return(p)
}
```

## 1.Import the data

```{r}
billboard <- read_csv("billboard.csv", 
                      col_types = cols(
                        url = col_skip(), 
                        song = col_skip(),
                        instance = col_skip()
                      ))
print(head(billboard,5))
```

```{r}
audio_features <- read_csv("audio_features.csv", 
                           col_types = cols( 
                             performer = col_skip(), 
                             song = col_skip(), 
                             spotify_track_preview_url = col_skip(),
                             spotify_track_explicit = col_skip(),
                             spotify_track_album = col_skip(),
                             spotify_track_id = col_skip()
                           ))
print(head(audio_features,5))
```

```{r}
billboard <- na.omit(billboard) # remove all NaN
audio_features <- as_tibble(audio_features) # translate in a modern and more flexible version
audio_features <- na.omit(audio_features) # remove all NaN

```

```{r}
full_data <- left_join(billboard, audio_features, by = "song_id") # merge datasets
```

## 2.Data exploration and cleaning

```{r}
glimpse(full_data)
```

```{r}
summary(full_data)
```

### 2.1 Missing Values

```{r}
na_counts <- count_na_per_column_df(full_data)
print(na_counts)
```

```{r}
full_data_clean <- full_data
full_data_clean$previous_week_position[is.na(full_data_clean$previous_week_position)] <- 0 # substitute NaN with 0
full_data_clean <- na.omit(full_data_clean) # remove all NaN

nrow(full_data_clean)
ncol(full_data_clean) 

```

### 2.2 Distributions

Distribution of numeric data

```{r}
numeric_data <- full_data_clean |> select(where(is.numeric))

numeric_data_long <- numeric_data |> pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# histograms distribution facet
ggplot(numeric_data_long, aes(x = value)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 25) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +  # 3 columns
  theme_minimal()
#ggsave("0_numeric_histogram_distribution.png", width = 12, height = 16)
```

QQ plot

```{r QQ-plots, cache=TRUE}
par(mfrow = c(2, 2))  # cambia a seconda del numero di variabili

for (colname in names(numeric_data)) {
  qqnorm(numeric_data[[colname]], main = paste("QQ Plot -", colname))
  qqline(numeric_data[[colname]])
}
```

Distribution of categorical data (due the fact there are too many unique values, we choose to filter and observe only the first 50 most frequent variables)

```{r}
categorical_data <- full_data_clean |> select(where(~ !is.numeric(.)))

categorical_data_long <- categorical_data |> pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# select top 50
categorical_data_top50 <- categorical_data_long |>
  group_by(variable, value) |>
  summarise(count = n(), .groups = "drop") |>
  group_by(variable) |>
  slice_max(order_by = count, n = 50)

# barplot distribution facet
ggplot(categorical_data_top50, aes(x = value, y = count)) +
  geom_col(fill = "steelblue", color = "white") + 
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  theme_minimal() +
  theme(#axis.text.x = element_text(angle = 45, hjust = 1))
    axis.text.x = element_blank() )
#ggsave("0_categorical_histogram_distribution.png", width = 12, height = 16)
```

### 2.3 Cleaning

```{r}
hist(full_data_clean$instrumentalness)
```

```{r}
hist(full_data_clean$time_signature)
```

```{r}
hist(full_data_clean$speechiness)
```

```{r}
hist(full_data_clean$spotify_track_duration_ms)
```

From the distribution of `instrumentalness` and `time signature` they seem to be not significant for our purpose. Then we decide to remove them.

```{r}
full_data_clean <- full_data_clean %>% select(-c(instrumentalness, time_signature))
```

### 2.4 Outliers

```{r}
# histograms distribution facet
ggplot(numeric_data_long, aes(x = value, y = "")) +
  geom_boxplot(fill = "steelblue") +
  facet_wrap(~ variable, scales = "free", ncol = 3) + 
  theme_minimal()
#ggsave("0_numeric_boxplot_distribution.png", width = 12, height = 16)
```

### 2.5 Correlations

```{r}
# compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# correlation heatmap
corrplot(cor_matrix, method = "color", 
         type = "upper", tl.col = "black", 
         #number.cex = 0.7, addCoef.col = "black",
         tl.cex = 0.8, col = colorRampPalette(c("purple", "white", "cyan"))(100))
#ggsave("0_numeric_heatmap.png", width = 12, height = 16)
```

## 3. Global Preprocessing

### 3.1 Aggregations

```{r}
# new features creation
billboard_summary <- billboard %>%
  group_by(song_id) %>%
  summarise(
    min_position = min(week_position),
    max_position = max(week_position),
    avg_position = mean(week_position),
    weeks_in_chart = n_distinct(week_id),
    final_position = last(week_position)
  )

full_data_aggregate <- merge(billboard_summary, audio_features, by = "song_id", all = FALSE)
full_data_aggregate <- full_data_aggregate %>% select(-c(instrumentalness, time_signature))

print(head(full_data_aggregate, 5))

```

```{r}
nrow(billboard)
nrow(billboard_summary)
nrow(audio_features)
nrow(full_data)
nrow(full_data_aggregate)
```

### 3.2 Add time columns

```{r}
full_data_clean <- full_data_clean |>
  select(colnames(full_data_clean),song_id,week_id)

full_data_clean$year <- as.numeric(sub(".*/.*/", "", full_data_clean$week_id))

head(full_data_clean, 5)
```

## 4 Main Question - Given the audio features of a new song, will it become popular?

### 4.1 Regression - predict the position on the billboard

#### Preprocessing

Remove unrelevant features

```{r}
regression_data <- full_data_aggregate %>% select(-c(song_id, min_position, max_position, avg_position, weeks_in_chart, final_position))
print(head(regression_data,5))
```

One hot encoding of the 15 most frequent genre

```{r}
regression_data
```


```{r}
regression_data <- extract_first_genre(regression_data)
regression_data_encoded <- process_and_encode_genre(regression_data, top_n = 15)
set.seed(45)
```

```{r}
regression_data$first_genre <- sapply(regression_data$first_genre, function(x) {
  # Rimuove i caratteri inutili
  cleaned <- gsub("\\[|\\]|'", "", x)
  # Divide in base alla virgola
  strsplit(cleaned, ",\\s*")[[1]][1]
})

ggplot(regression_data, aes(x = first_genre)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of the first genre",
       x = "Genre",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
As you can see a lot of the genres that are present in the dataset are not really used that much, so we have decided to take the first 15.

Scaling

```{r}
regression_data_encoded_scaled <- robustHD::standardize(regression_data_encoded |> select(-spotify_track_popularity))

regression_data_encoded_scaled['spotify_track_popularity'] = regression_data$spotify_track_popularity

regression_data_encoded_scaled
```

#### Regressions

```{r big-model, cache=TRUE}
library(tidymodels)
library(furrr)
set.seed(45)

plan(multisession, workers = 6)

# 1. Split
split <- initial_split(regression_data_encoded_scaled, prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)

# 2. Recipe
recipe_spec <- recipe(spotify_track_popularity ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

# 3. Resampling (non usato direttamente, ma mantenuto)
cv_folds <- vfold_cv(train_data, v = 5)

# 4. Modelli con parametri fissi
models <- list(
  rf = rand_forest(mtry = 5, trees = 500) %>%
    set_engine("randomForest") %>% set_mode("regression"),

  xgb = boost_tree(trees = 1000, tree_depth = 6, learn_rate = 0.1) %>%
    set_engine("xgboost") %>% set_mode("regression"),

  knn = nearest_neighbor(neighbors = 7) %>%
    set_engine("kknn") %>% set_mode("regression")
)

# 5. Workflow per ciascun modello
workflows_list <- purrr::imap(models, function(model, name) {
  workflow() %>%
    add_model(model) %>%
    add_recipe(recipe_spec)
})

# 6. Fit parallelo
final_models <- furrr::future_map(workflows_list, function(wf) {
  library(tidymodels)  # necessario dentro il worker
  fit(wf, data = train_data)
})

# 7. Predizione + metriche parallele
results <- furrr::future_map(final_models, ~ {
  predict(.x, test_data) %>%
    bind_cols(test_data) %>%
    metrics(truth = spotify_track_popularity, estimate = .pred)
})

# 8. Riassunto risultati
results_df <- bind_rows(results, .id = "model") %>%
  filter(.metric == "rmse") %>%
  arrange(.estimate)

print(results_df)

# Torna a sequenziale
plan(sequential)
```

```{r}
all_predictions <- purrr::imap(final_models, ~ {
  predict(.x, test_data) |>
    bind_cols(test_data) |>
    mutate(model = .y)
}) |> bind_rows()

```

```{r}
min_val <- min(c(all_predictions$spotify_track_popularity, all_predictions$.pred))
max_val <- max(c(all_predictions$spotify_track_popularity, all_predictions$.pred))

ggplot(all_predictions, aes(x = spotify_track_popularity, y = .pred)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~model, scales = "free") +
  labs(
    title = "Predictions vs Real Values for Models",
    x = "Real Values (spotify_track_popularity)",
    y = "Predictions"
  ) +
  theme_minimal() +
  coord_cartesian(xlim = c(min_val, max_val), ylim = c(min_val, max_val))

```
As you can clearly state from the Regressions above, the model is not predicting well the final popularity of the song. All three of the model used in this task have a low performance. The lowest RMSE comes from the Random Forest regressor with a value of 16.8 and is quite high. We want to predict a value that goes from 0 to 100 and with an RMSE of almost 16.8 we are not precise enough. If the true value of the popularity is 50 that means that our model could predict a value that goes from 31.2 to 66.8 and this is not precise enough for us. Is there a way to improove the model performances reducing the RMSE?

First let's try to understand wich features are the most imporant for our model.

features importance

```{r}
# VIP plot list
vip_plots <- list()

# Random Forest
rf_fit <- extract_fit_parsnip(final_models$rf)
if (inherits(rf_fit$fit, "randomForest")) {
  vip_plots$rf <- vip(rf_fit$fit, num_features = 15) + ggtitle("Random Forest")
}

# XGBoost
xgb_fit <- extract_fit_parsnip(final_models$xgb)
if (inherits(xgb_fit$fit, "xgb.Booster")) {
  vip_plots$xgb <- vip(xgb_fit$fit, num_features = 15) + ggtitle("XGBoost")
}

# visualize plots
library(patchwork)

# show them
if (length(vip_plots) == 2) {
  vip_plots$rf + vip_plots$xgb
} else {
  # if only one is possible
  vip_plots[[1]]
}
```
The models have chosen the same features to characterize as the most important ones. We have Loudness in firs place and then the duration of the track in second place, after this common grounds the models use the same features but in a different order. 

One thing that we could try to do to improove the performances of our model is trying to bin the target popularity tor see if this process help the model.

We have decided to bin the popularity in 4 different features each of 25 points with the labels of low, medium, high, very high.

### 4.2 Classification- predict which level of popularity

#### Preprocessing

Remove unrelevant features

```{r}
classification_data <- full_data_aggregate %>% select(-c( song_id, min_position, max_position, avg_position, weeks_in_chart, final_position))
print(head(classification_data,5))
```

Encode popularity

```{r}
classification_data <- encode_spotify_popularity(classification_data)

classification_data <- classification_data |>
  filter(!is.na(spotify_track_popularity)) 
```

Encode genre

```{r}
classification_data <- extract_first_genre(classification_data)
classification_data <- process_and_encode_genre(classification_data, top_n = 15)
```

```{r}
print(head(classification_data,5))
```

Analyse the correlations

```{r}
# select only numerical data
numeric_data_clss <- classification_data |>
  select(where(is.numeric))

# compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# correlation heatmap
corrplot(cor_matrix, method = "color", 
         type = "upper", tl.col = "black", 
         #number.cex = 0.7, addCoef.col = "black",
         tl.cex = 0.8, col = colorRampPalette(c("purple", "white", "cyan"))(100))

#ggsave("4_regression_heatmap.png", width = 12, height = 16)
```

Target Distribution

```{r}
ggplot(classification_data, aes(x = spotify_track_popularity)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Popularity Distribution",
       x = "Spotify Track Popularity",
       y = "Counter") +
  theme_minimal()
```
We can see that the target values are not equally rerpresented, this is a problem for our modell because will have an arder time in predicting songs with very high popularity.


split

```{r}
set.seed(45)
spotify_train_test <- classification_data |> initial_split(prop =  0.7)
spotify_train <- training(spotify_train_test) 
spotify_test  <- testing(spotify_train_test)
```

#### Classification

Random forest

```{r}
rand_forest_spec <- 
  rand_forest(mtry = 5, trees = 200) |>  
  set_engine('randomForest') |> 
  set_mode('classification')

rf_recipe = recipe(spotify_track_popularity~.,
                   data=spotify_train)

rf_wf <- workflow() |> 
  add_recipe(rf_recipe) |> 
  add_model(rand_forest_spec)

rf_wf_fit <- rf_wf |> 
  fit(spotify_train)

rf_wf_fit
```

```{r}
spotify_test_rf <- rf_wf_fit |> 
  augment(spotify_test)

# confusion matrix
cm_rF <- spotify_test_rf |> 
  conf_mat(truth = spotify_track_popularity,
           estimate = .pred_class)

# visualize
plot_conf_matrix(cm_rF, title = "Confusion Matrix - Classification Random Forest")

```

```{r}
head(summary(cm_rF),4)
```
From this evaluation we can see that binning the target value and transforming this problem in a classification problem did not help much. An accuracy just above 53% percent is still a very low performance.


```{r}
rf_wf_fit |> 
  extract_fit_engine() |> 
  vip()
```
From the feature importance of the model above we can claearly state that the most important features are still the same as before.

I idea that came into our head while working at this project was that we are mostly interested in predicting the popularity of the new songs that get produced today, not the ones that were published or famous in the 1960s. Maybe if we select only the last 10 years of the dataset we will have an higher accuracy because the taste of the peoples hearing the music will have less changes. Afterall the music that was the mos poupular in the 60s is not the same one that we listen to today.

> Is the music changed over time?
  
## 5. Taking only the last 10 years of data, <<<<<<<<the model swill improve / possibile stimare meglio>>>>>>>>>>>>>?
### 5.1 Preprocessing
  Remove unrelevant features
```{r}
data_filtered <- full_data_clean |> select(-c(week_id, week_position, previous_week_position, peak_position, weeks_on_chart))
```

convert milliseconds in minutes
```{r}
data_filtered$spotify_track_duration_min <- data_filtered$spotify_track_duration_ms/60000
data_filtered <- data_filtered |> select(-c(spotify_track_duration_ms))
```

Filter the song after 2000
```{r}
bound_year = 2010

data_filtered_after <- data_filtered[data_filtered$year >= bound_year & data_filtered$year < 2021, ]
data_filtered_before <- data_filtered[data_filtered$year < bound_year, ]
```

encode the most frequent 15 genres for each period
```{r}
data_filtered_after_encoded <- extract_first_genre(data_filtered_after) 
data_filtered_after_encoded <- process_and_encode_genre(data_filtered_after_encoded, top_n = 15)

data_filtered_before_encoded <- extract_first_genre(data_filtered_before) 
data_filtered_before_encoded <- process_and_encode_genre(data_filtered_before_encoded, top_n = 15)
```

remove duplicates
```{r}
data_filtered_after_encoded$year <- NULL
data_filtered_after_encoded <- unique(data_filtered_after_encoded)
data_filtered_before_encoded <- unique(data_filtered_before_encoded)
```

```{r}
head(data_filtered_after_encoded,5)
```

### 5.2 Regression after 2010
#### Preprocessing
```{r}
regression_data_after <- data_filtered_after_encoded |> select(-c(performer, song_id))
regression_data_after
```

Scale the features
```{r}
#regression_data_after <- robustHD::standardize(regression_data_after |> select(-c( year, spotify_track_popularity)))
regression_data_after$spotify_track_popularity <- data_filtered_after_encoded$spotify_track_popularity
regression_data_after <- unique(regression_data_after)

#regression_data_after$year <- data_filtered_after_encoded$ye
head(regression_data)
```

#### Regressions

```{r}
# compute correlation matrix
cor_matrix <- cor(regression_data_after, use = "complete.obs")

# correlation heatmap
corrplot(cor_matrix, method = "color", 
         type = "upper", tl.col = "black", 
         #number.cex = 0.7, addCoef.col = "black",
         tl.cex = 0.8, col = colorRampPalette(c("purple", "white", "cyan"))(100))
```
From this correlation matrix we can see that the features on wich we work are not much correlated with each other, We have only one strong correlation between energy and loudness but that should not be a problem for the models that we are going to use.

```{r small-model, cache=True}
set.seed(45)
library(tidymodels)
library(furrr)

plan(multisession, workers = 6)

# 1. Split
split <- initial_split(regression_data_after, prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)

# 2. Recipe
recipe_spec <- recipe(spotify_track_popularity ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

# 3. Resampling (non usato qui)
cv_folds <- vfold_cv(train_data, v = 5)

# 4. Modelli con parametri fissi
models <- list(
  rf = rand_forest(mtry = 5, trees = 500) %>%
    set_engine("randomForest") %>% set_mode("regression"),

  xgb = boost_tree(trees = 1000, tree_depth = 6, learn_rate = 0.1) %>%
    set_engine("xgboost") %>% set_mode("regression"),

  knn = nearest_neighbor(neighbors = 7) %>%
    set_engine("kknn") %>% set_mode("regression")
)

# 5. Workflows
workflows_list <- purrr::imap(models, function(model, name) {
  workflow() %>%
    add_model(model) %>%
    add_recipe(recipe_spec)
})

# 6. Fit parallelo
final_models <- furrr::future_map(workflows_list, function(wf) {
  library(tidymodels)  # necessario dentro il worker
  fit(wf, data = train_data)
})

# 7. Valuta su test set in parallelo
results <- furrr::future_map(final_models, ~ {
  predict(.x, test_data) %>%
    bind_cols(test_data) %>%
    metrics(truth = spotify_track_popularity, estimate = .pred)
})

# 8. Riassunto RMSE
results_df <- bind_rows(results, .id = "model") %>%
  filter(.metric == "rmse") %>%
  arrange(.estimate)

print(results_df)

# (opzionale) chiudi il cluster
plan(sequential)
```

```{r}
# compute correlation matrix
cor_matrix <- cor(regression_data_after, use = "complete.obs")

# correlation heatmap
corrplot(cor_matrix, method = "color", 
         type = "upper", tl.col = "black", 
         #number.cex = 0.7, addCoef.col = "black",
         tl.cex = 0.8, col = colorRampPalette(c("purple", "white", "cyan"))(100))

#ggsave("4_regression_heatmap.png", width = 12, height = 16)
```


```{r}
all_predictions <- purrr::imap(final_models, ~ {
  predict(.x, test_data) |>
    bind_cols(test_data) |>
    mutate(model = .y)
}) |> bind_rows()

```

```{r}
min_val <- min(c(all_predictions$spotify_track_popularity, all_predictions$.pred))
max_val <- max(c(all_predictions$spotify_track_popularity, all_predictions$.pred))

ggplot(all_predictions, aes(x = spotify_track_popularity, y = .pred)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~model, scales = "free") +
  labs(
    title = "Predictions vs Real Values for Models",
    x = "Real Values (spotify_track_popularity)",
    y = "Predictions"
  ) +
  theme_minimal() +
  coord_cartesian(xlim = c(min_val, max_val), ylim = c(min_val, max_val))

```
The new models trained on only the final section of the data we have are more precise than the general ones. This is a good news because it means that we are able to produce more precise prediction over the songs of the last 10ish years. Now the lowest RMSE is around 10 that for us is a good improovement. With this new model if we want to predict 50 as popularity for a particular song we will end up with values ranging from 40 to 60 that is a more acceptable result.

features importance
```{r}
# VIP plot list
vip_plots <- list()

# Random Forest
rf_fit <- extract_fit_parsnip(final_models$rf)
if (inherits(rf_fit$fit, "randomForest")) {
  vip_plots$rf <- vip(rf_fit$fit, num_features = 15) + ggtitle("Random Forest")
}

# XGBoost
xgb_fit <- extract_fit_parsnip(final_models$xgb)
if (inherits(xgb_fit$fit, "xgb.Booster")) {
  vip_plots$xgb <- vip(xgb_fit$fit, num_features = 15) + ggtitle("XGBoost")
}

# visualize plots
library(patchwork)

# show them
if (length(vip_plots) == 2) {
  vip_plots$rf + vip_plots$xgb
} else {
  # if only one is possible
  vip_plots[[1]]
}
```
This new model uses the same features of the initial one so probably would be interesting to explore the behaviour of this data to see if there are patterns that we can recognize

```{r}
loudness_summary <- full_data_clean %>%
  group_by(year) %>%
  summarise(
    mean_loudness = mean(loudness, na.rm = TRUE),
    sd_loudness = 3 * sd(loudness, na.rm = TRUE),
    n = n(),
    se = sd_loudness / sqrt(n)  # standard error
  )

ggplot(loudness_summary, aes(x = year, y = mean_loudness)) +
  geom_line(color = "steelblue", size = 1) +
  geom_ribbon(aes(ymin = mean_loudness - se, ymax = mean_loudness + se), 
              fill = "lightblue", alpha = 0.4) +
  labs(title = "Loudness Through the Years",
       x = "Year", y = "Average Loudness (dB)") +
  theme_minimal()
```
Well for the loudness we can see that after 1990 the loudness inside song had a big rise, this is due to the fact that loud songs are easier to listen with headphones. But one thing that we can notice is that the loudness of the last years is much more different than the loudness before the 1990, this probably helped the model in deciding how to predict the final popularity of a song.

```{r}

loudness_summary <- full_data_clean %>%
  group_by(year) %>%
  summarise(
    mean_loudness = mean(spotify_track_duration_ms, na.rm = TRUE),
    sd_loudness = 3 * sd(spotify_track_duration_ms, na.rm = TRUE),
    n = n(),
    se = sd_loudness / sqrt(n)  # standard error
  )

ggplot(loudness_summary, aes(x = year, y = mean_loudness)) +
  geom_line(color = "steelblue", size = 1) +
  geom_ribbon(aes(ymin = mean_loudness - se, ymax = mean_loudness + se), 
              fill = "lightblue", alpha = 0.4) +
  labs(title = "Track duration over the Years",
       x = "Year", y = "track duration (ms)") +
  theme_minimal()

```

From this graph we can see that the duration of a song between 2010 and 2021 is quite different from the ones before 2010. We can also notice that in recent history were first short, than they became longher and now are also decreasing in time, a really interesting pattern.

```{r}

loudness_summary <- full_data_clean %>%
  group_by(year) %>%
  summarise(
    mean_loudness = mean(energy, na.rm = TRUE),
    sd_loudness = 3 * sd(energy, na.rm = TRUE),
    n = n(),
    se = sd_loudness / sqrt(n)  # standard error
  )

ggplot(loudness_summary, aes(x = year, y = mean_loudness)) +
  geom_line(color = "steelblue", size = 1) +
  geom_ribbon(aes(ymin = mean_loudness - se, ymax = mean_loudness + se), 
              fill = "lightblue", alpha = 0.4) +
  labs(title = "Acousticness over the Years",
       x = "Year", y = "Acousticness") +
  theme_minimal()

```
Also here we see that acousticness between 2010 and 2021 is much more different than the past years.

Overall we can see that since the characteristics of music in different years is changed in many ways and not by little, so it make sense to create different models for the different periods of time to approximate better the taste of the people living in a particular period. It would probably be interesting to do this reasoning by country.

## 6. Have songs from the last ten years become louder than before? We chose to perform this hypothesis test on the loudness because we noticed it was the most important feature among all the audio features of the songs.

### 6.1 Hypothesis test
#### Preprocessing

Remove unrelevant features
```{r}
# selecting only the numeric features
hypothesis_data_after <- data_filtered_after_encoded |> select(where(is.numeric))
hypothesis_data_before <- data_filtered_before_encoded |> select(where(is.numeric))
```

Analyse distribution
```{r}
#selecting all the features that are not related to the genre
numeric_data_after_no_genre <- hypothesis_data_after |>
  select(1:(ncol(hypothesis_data_after) - 11))
par(mfrow = c(2,2))

numeric_data_before_no_genre <- hypothesis_data_before |>
  select(1:(ncol(hypothesis_data_before) - 11))
```

*After 2000*
```{r}
#plotting the QQ plots to show the distribution of the dataset
par(mfrow = c(2,2))
for (colname in names(numeric_data_after_no_genre)) {
  qqnorm(numeric_data_after_no_genre[[colname]], main = paste("QQ Plot -", colname))
  qqline(numeric_data_after_no_genre[[colname]])
}
```
*Before 2000*
```{r}
#plotting the QQ plots to show the distribution of the dataset
par(mfrow = c(2,2))
for (colname in names(numeric_data_before_no_genre)) {
  qqnorm(numeric_data_before_no_genre[[colname]], main = paste("QQ Plot -", colname))
  qqline(numeric_data_before_no_genre[[colname]])
}
```

Based on the QQ plots for the periods before and after 2015, none of the feature distributions appear to be normal. Therefore, to compare the two distributions, we should use the Wilcoxon test.

#### Hypothesis Test
```{r}
#Doing a wilcox test based on the loudness of the songs
wilcox.test(
  hypothesis_data_after$spotify_track_duration_min,
  hypothesis_data_before$spotify_track_duration_min,
  alternative = "two.sided"
)

#filtering the datasets with only the loudness feature
loudness_after <- hypothesis_data_after$spotify_track_duration_min
loudness_before <- hypothesis_data_before$spotify_track_duration_min


plot(density(loudness_after), col = "blue", lwd = 2,
     main = "Spotify track duration min Density: After 2010 vs Before 2010", xlab = "duration")
lines(density(loudness_before), col = "red", lwd = 2)

legend("topright", legend = c("After 2010", "Before 2010"), col = c("blue", "red"), lwd = 2)
```
The p-value from this test is very close to the acceptance of H1 but is just to high for us. This could be due to an unlucky case. Anyway we can see that the duration of the tracks after 2010 are more focused on a single mean just under 4 minutes, while the ones before are a bit more spread out between 2 and 4 minutes. Anyway we dont see much difference in the two distributions.


```{r}
#Doing a wilcox test based on the loudness of the songs
wilcox.test(
  hypothesis_data_after$energy,
  hypothesis_data_before$energy,
  alternative = "two.sided"
)

#filtering the datasets with only the loudness feature
loudness_after <- hypothesis_data_after$energy
loudness_before <- hypothesis_data_before$energy


plot(density(loudness_after), col = "blue", lwd = 2,
     main = "Energy Density: After 2010 vs Before 2010", xlab = "energy")
lines(density(loudness_before), col = "red", lwd = 2)

legend("topright", legend = c("After 2010", "Before 2010"), col = c("blue", "red"), lwd = 2)
```
As we can see from the two distributions, songs released after 2010 are almost the energy dense than the ones released before.


```{r}
#Doing a wilcox test based on the loudness of the songs
wilcox.test(
  hypothesis_data_after$loudness,
  hypothesis_data_before$loudness,
  alternative = "two.sided"
)

#filtering the datasets with only the loudness feature
loudness_after <- hypothesis_data_after$loudness
loudness_before <- hypothesis_data_before$loudness


plot(density(loudness_after), col = "blue", lwd = 2,
     main = "Spotify track duration min Density: After 2010 vs Before 2010", xlab = "duration")
lines(density(loudness_before), col = "red", lwd = 2)

legend("topright", legend = c("After 2010", "Before 2010"), col = c("blue", "red"), lwd = 2)
```


As we can see from the two distributions, songs released after 2010 are louder than those released before. Why is that? This is due to a big movement in the early 90ntees that wanted to make songs louder because they are easier to listen to.


## Question 2 - As a manager, should I continue to invest in this song given its statistics? Will it remain popular over time?

### Regression - predict the position on the Billboard

#### Preprocessing

Remove unrelevant features

```{r}
regression_data <- full_data_aggregate %>% select(-c(song_id))
print(head(regression_data,5))
```

one hot encoding of the most 15 important genres

```{r}
regression_data <- extract_first_genre(regression_data)
regression_data_encoded <- process_and_encode_genre(regression_data, top_n = 15)
```

```{r}
print(head(regression_data_encoded, 5))
```

#### Regression

```{r}
set.seed(45)
split <- initial_split(head(regression_data_encoded,100), prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)

# recipe
recipe_spec <- recipe(final_position ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

task = 'regression'

# models
models <- list(
  rf = rand_forest(mtry = 5) %>% set_engine("randomForest") %>% set_mode(task),
  #lm = linear_reg() %>% set_engine("lm") %>% set_mode(task),
  xgb = boost_tree(trees = 100, tree_depth = 10) %>% set_engine("xgboost") %>% set_mode(task),
  knn = nearest_neighbor(neighbors = 5) %>% set_engine("kknn") %>% set_mode(task)
)

# workflow
workflows_list <- purrr::map(models, ~ workflow() %>% add_model(.x) %>% add_recipe(recipe_spec))

# train models
fitted_models <- purrr::map(workflows_list, ~ fit(.x, data = train_data))

# evaluation
results <- purrr::map(fitted_models, ~ {
  predict(.x, test_data) %>%
    bind_cols(test_data) %>%
    metrics(truth = final_position, estimate = .pred)
})

# results
results_df <- bind_rows(results, .id = "model")

results_df %>%
  filter(.metric == "rmse") %>%
  arrange(.estimate)

print(results_df)
```


```{r}
# VIP plot list
vip_plots <- list()

# Random Forest
rf_fit <- extract_fit_parsnip(fitted_models$rf)
if (inherits(rf_fit$fit, "randomForest")) {
  vip_plots$rf <- vip(rf_fit$fit, num_features = 15) + ggtitle("Random Forest - Variable Importance")
}

# XGBoost
xgb_fit <- extract_fit_parsnip(fitted_models$xgb)
if (inherits(xgb_fit$fit, "xgb.Booster")) {
  vip_plots$xgb <- vip(xgb_fit$fit, num_features = 15) + ggtitle("XGBoost - Variable Importance")
}

# visualize plots
library(patchwork)

# show them
if (length(vip_plots) == 2) {
  vip_plots$rf + vip_plots$xgb
} else {
  # if only one is possible
  vip_plots[[1]]
}
```

### Classification- predict whether a sign will be in the top 30

#### Preprocessing

```{r}
classification_data <- full_data_aggregate |> select(-c( song_id))
print(head(classification_data,5))
```

Encode popularity

```{r}
classification_data <- encode_spotify_popularity(classification_data)

classification_data <- classification_data |>
  filter(!is.na(spotify_track_popularity)) 
```

Encode genre

```{r}
classification_data <- extract_first_genre(classification_data)
classification_data <- process_and_encode_genre(classification_data, top_n = 5)
```

```{r}
#print number of NaN for each column
count_na_per_column_df(classification_data)
```

```{r}
#create the new variable top30: whether a song is in the top 30
classification_data <- classification_data |>
  mutate(top30 = ifelse(final_position <= 30, 1, 0))
# remove final_position
classification_data <- classification_data |> select(-c( final_position))
print(head(classification_data,5))
```

```{r}

# select only numerical data
numeric_data_clss <- numeric_data_clss |>
  select(where(is.numeric))

# compute correlation matrix
cor_matrix <- cor(numeric_data_clss, use = "complete.obs")

# correlation heatmap
corrplot(cor_matrix, method = "color", 
         type = "upper", tl.col = "black", 
         #number.cex = 0.7, addCoef.col = "black",
         tl.cex = 0.8, col = colorRampPalette(c("purple", "white", "cyan"))(100))

#ggsave("4_regression_heatmap.png", width = 12, height = 16)
```
```{r}
hist(classification_data$top30)
```

#### Classification

```{r}
set.seed(42)
classification_data$top30 <- as.factor(classification_data$top30)
spotify_train_test <- classification_data |> initial_split(prop =  0.7)
spotify_train <- training(spotify_train_test) 
spotify_test  <- testing(spotify_train_test)
```

```{r}
rand_forest_spec <- 
  rand_forest(mtry = 5, trees = 200) |>  
  set_engine('randomForest') |> 
  set_mode('classification')

rf_recipe = recipe(top30~.,
                   data=spotify_train)

rf_wf <- workflow() |> 
  add_recipe(rf_recipe) |> 
  add_model(rand_forest_spec)

rf_wf_fit <- rf_wf |> 
  fit(spotify_train)

rf_wf_fit
```

```{r}
spotify_test_rf <- rf_wf_fit |> 
  augment(spotify_test)

# confusion matrix
cm_rF <- spotify_test_rf |> 
  conf_mat(truth = top30,
           estimate = .pred_class)

# visualize
plot_conf_matrix(cm_rF, title = "Confusion Matrix - Classification Random Forest")
```

```{r}
head(summary(cm_rF),4)
```

```{r}
roc_rF<-roc_curve(spotify_test_rf,
          top30,
          .pred_0,event_level = "first")
roc_rF |> autoplot()
```

```{r}
rf_wf_fit |> 
  extract_fit_engine() |> 
  vip()
```


