---
title: "R Notebook"
output: html_notebook
---

# Spotify Project

In this project we will analyse the evolution of the music, from 1956 until now.

-   First we start to explore whether is possible to predict the popularity of a song from its audio features

-   In second

link to the **dataset**: [jonhthgeek-DataSetSpotify](https://github.com/rfordatascience/tidytuesday/blob/main/data/2021/2021-09-14/readme.md)

**Authors**: [Bernacchia Alessia](https://github.com/AlessiaBernacchia), [Pioda Tommaso](https://github.com/Thetommigun432), [Villani Giacomo](https://github.com/DownToTheGround)

```{r}
# samples for regression
n_samples_to_consider = 100
```


```{r}
library(vip)
library(readr)
library(dplyr)
library(tidymodels)
library(ggplot2)
library(rsample)
library(tidyr)
library(themis)
library(stringr)
library(purrr)
library(caret)
library(kknn)

#install.packages("xgboost")
library(parsnip)
library(randomForest)
library(kknn)
library(xgboost)
library(finetune)
library(doParallel)
library(robustHD)
library(parsnip)

# for exploration
library(corrplot)
library(ggcorrplot)

# for classification
library(pROC)
```

*Functions*:

```{r}
# count NaN for each column
count_na_per_column <- function(data) {
  sapply(data, function(col) sum(is.na(col)))}

count_na_per_column_df <- function(data) {
  na_counts <- sapply(data, function(col) sum(is.na(col)))
  na_counts_df <- data.frame(
    feature = names(na_counts),
    na_count = na_counts
  )
  na_counts_df %>% arrange(desc(na_count))
}

# extract first genre for the spotify genre
extract_first_genre <- function(df, genre_column = "spotify_genre") {
  # check if column exists
  if (!genre_column %in% colnames(df)) {
    stop(paste("La colonna", genre_column, "non esiste nel dataframe"))
  }
  
  # removes useless not required characters
  df[[genre_column]] <- gsub("\\[|\\]|'", "", df[[genre_column]])
  
  # create a list, with div = ,
  genre_list <- strsplit(df[[genre_column]], split = ",")
  
  # extract only the first genre for each line or rename "others"
  df$first_genre <- sapply(genre_list, function(x) {
    x <- trimws(x)  # remove blanks
    if (length(x) == 0 || is.na(x[1]) || x[1] == "") {
      return("others")
    } else {
      return(x[1])
    }
  })
  
  # convert to categorical value
  df$first_genre <- as.factor(df$first_genre)
  
  # output
  return(df)
}

# process and encode the genre column
process_and_encode_genre <- function(df, genre_column = "first_genre", top_n = 15) {
  # check if column exists
  if (!genre_column %in% colnames(df)) {
    stop(paste("La colonna", genre_column, "non esiste nel dataframe"))
  }
  
  # count the frequency
  genre_counts <- table(df[[genre_column]])
  
  # put them in decrescent order
  genre_counts_sorted <- sort(genre_counts, decreasing = TRUE)
  
  # take top N genres
  top_genres <- names(genre_counts_sorted)[1:min(top_n, length(genre_counts_sorted))]
  
  # rename not in top N into 'others'
  df[[genre_column]] <- ifelse(
    df[[genre_column]] %in% top_genres,
    df[[genre_column]],
    "others"
  )
  
  # convert
  df[[genre_column]] <- as.factor(df[[genre_column]])
  
  # remove original column
  if ("spotify_genre" %in% colnames(df)) {
    df <- subset(df, select = -c(spotify_genre))
  }
  
  # create dummy variables (one-hot encoding)
  dummy <- dummyVars(paste("~", genre_column), data = df)
  dummy_df <- data.frame(predict(dummy, newdata = df))
  
  final_df <- bind_cols(
    df %>% select(-all_of(genre_column)),
    dummy_df
  )
  
  # output
  return(final_df)
}

# encode popularity column
encode_spotify_popularity <- function(df, column = "spotify_track_popularity") {
  # check if column exists
  if (!column %in% colnames(df)) {
    stop(paste("La colonna", column, "non esiste nel dataframe"))
  }
  
  categorize_popularity <- function(x) {
    if (is.na(x)) return(NA)
    if (x <= 25) {
      return("low")
    } else if (x <= 50) {
      return("medium")
    } else if (x <= 75) {
      return("high")
    } else if (x <= 100) {
      return("very high")
    } else {
      return(NA)  # out of range
    }
  }
  
  df[[column]] <- sapply(df[[column]], categorize_popularity)
  
  # convert in ordered factor
  df[[column]] <- factor(df[[column]], 
                         levels = c("low", "medium", "high", "very high"), 
                         ordered = TRUE)
  
  # output
  return(df)
}

# plots the confusion matrix
plot_conf_matrix <- function(conf_matrix_object, title = "Confusion Matrix") {
  
  # converts confusion matrix into dataframe
  cm_df <- as.data.frame(conf_matrix_object$table)
  
  # find columns names (Truth, Prediction, Freq)
  colnames(cm_df) <- tolower(colnames(cm_df)) # rende tutto minuscolo
  
  # rename them
  names(cm_df)[names(cm_df) %in% c("reference", "truth")] <- "Actual"
  names(cm_df)[names(cm_df) == "prediction"] <- "Predicted"
  names(cm_df)[names(cm_df) %in% c("freq", "n")] <- "Freq"
  
  # plot
  p <- ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Freq)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = Freq), size = 5, color = "black") +
    scale_fill_gradient(low = "white", high = "steelblue") +
    labs(title = title,
         x = "Predicted Class",
         y = "Actual Class",
         fill = "Count") +
    theme_minimal(base_size = 14) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  return(p)
}
```

## 1.Import the data

```{r}
billboard <- read_csv("billboard.csv", 
                      col_types = cols(
                        url = col_skip(), 
                        song = col_skip(),
                        instance = col_skip()
                      ))
print(head(billboard,5))
```

```{r}
audio_features <- read_csv("audio_features.csv", 
                           col_types = cols( 
                             performer = col_skip(), 
                             song = col_skip(), 
                             spotify_track_preview_url = col_skip(),
                             spotify_track_explicit = col_skip(),
                             spotify_track_album = col_skip(),
                             spotify_track_id = col_skip()
                           ))
print(head(audio_features,5))
```

```{r}
billboard <- na.omit(billboard) # remove all NaN
audio_features <- as_tibble(audio_features) # translate in a modern and more flexible version
audio_features <- na.omit(audio_features) # remove all NaN

```

```{r}
full_data <- left_join(billboard, audio_features, by = "song_id") # merge datasets
```

## 2.Data exploration and cleaning

```{r}
glimpse(full_data)
```

```{r}
summary(full_data)
```

### 2.1 Missing Values

```{r}
na_counts <- count_na_per_column_df(full_data)
print(na_counts)
```

```{r}
full_data_clean <- full_data
full_data_clean$previous_week_position[is.na(full_data_clean$previous_week_position)] <- 0 # substitute NaN with 0
full_data_clean <- na.omit(full_data_clean) # remove all NaN

nrow(full_data_clean)
ncol(full_data_clean) 

```

### 2.2 Distributions

Distribution of numeric data

```{r}
numeric_data <- full_data_clean |> select(where(is.numeric))

numeric_data_long <- numeric_data |> pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# histograms distribution facet
ggplot(numeric_data_long, aes(x = value)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 25) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +  # 3 columns
  theme_minimal()
#ggsave("0_numeric_histogram_distribution.png", width = 12, height = 16)
```

QQ plot

```{r QQ-plots, cache=TRUE}
par(mfrow = c(2, 2))  # cambia a seconda del numero di variabili

for (colname in names(numeric_data)) {
  qqnorm(numeric_data[[colname]], main = paste("QQ Plot -", colname))
  qqline(numeric_data[[colname]])
}
```

Distribution of categorical data (due the fact there are too many unique values, we choose to filter and observe only the first 50 most frequent variables)

```{r}
categorical_data <- full_data_clean |> select(where(~ !is.numeric(.)))

categorical_data_long <- categorical_data |> pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# select top 50
categorical_data_top50 <- categorical_data_long |>
  group_by(variable, value) |>
  summarise(count = n(), .groups = "drop") |>
  group_by(variable) |>
  slice_max(order_by = count, n = 50)

# barplot distribution facet
ggplot(categorical_data_top50, aes(x = value, y = count)) +
  geom_col(fill = "steelblue", color = "white") + 
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  theme_minimal() +
  theme(#axis.text.x = element_text(angle = 45, hjust = 1))
    axis.text.x = element_blank() )
#ggsave("0_categorical_histogram_distribution.png", width = 12, height = 16)
```

### 2.3 Cleaning

```{r}
hist(full_data_clean$instrumentalness)
```

```{r}
hist(full_data_clean$time_signature)
```

```{r}
hist(full_data_clean$speechiness)
```

```{r}
hist(full_data_clean$spotify_track_duration_ms)
```

From the distribution of `instrumentalness` and `time signature` they seem to be not significant for our purpose. Then we decide to remove them.

```{r}
full_data_clean <- full_data_clean %>% select(-c(instrumentalness, time_signature))
```

### 2.4 Outliers

```{r}
# histograms distribution facet
ggplot(numeric_data_long, aes(x = value, y = "")) +
  geom_boxplot(fill = "steelblue") +
  facet_wrap(~ variable, scales = "free", ncol = 3) + 
  theme_minimal()
#ggsave("0_numeric_boxplot_distribution.png", width = 12, height = 16)
```

### 2.5 Correlations

```{r}
# compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# correlation heatmap
corrplot(cor_matrix, method = "color", 
         type = "upper", tl.col = "black", 
         #number.cex = 0.7, addCoef.col = "black",
         tl.cex = 0.8, col = colorRampPalette(c("purple", "white", "cyan"))(100))
#ggsave("0_numeric_heatmap.png", width = 12, height = 16)
```

## 3. Global Preprocessing

### 3.1 Aggregations

```{r}
# new features creation
billboard_summary <- billboard %>%
  group_by(song_id) %>%
  summarise(
    min_position = min(week_position),
    max_position = max(week_position),
    avg_position = mean(week_position),
    weeks_in_chart = n_distinct(week_id),
    final_position = last(week_position)
  )

full_data_aggregate <- merge(billboard_summary, audio_features, by = "song_id", all = FALSE)
full_data_aggregate <- full_data_aggregate %>% select(-c(instrumentalness, time_signature))

print(head(full_data_aggregate, 5))

```

```{r}
nrow(billboard)
nrow(billboard_summary)
nrow(audio_features)
nrow(full_data)
nrow(full_data_aggregate)
```

### 3.2 Add time columns

```{r}
full_data_clean <- full_data_clean |>
  select(colnames(full_data_clean),song_id,week_id)

full_data_clean$year <- as.numeric(sub(".*/.*/", "", full_data_clean$week_id))

head(full_data_clean, 5)
```

## 4 Main Question - Given the audio features of a new song, will it become popular?

### 4.1 Regression - predict the position on the billboard

#### Preprocessing

Remove unrelevant features

```{r}
regression_data <- full_data_aggregate %>% select(-c(song_id, min_position, max_position, avg_position, weeks_in_chart, final_position))
print(head(regression_data,5))
```

One hot encoding of the 15 most frequent genre

```{r}
regression_data <- extract_first_genre(regression_data)
regression_data_encoded <- process_and_encode_genre(regression_data, top_n = 15)
set.seed(45)
```

Scaling

```{r}
regression_data_encoded_scaled <- robustHD::standardize(regression_data_encoded |> select(-spotify_track_popularity))

regression_data_encoded_scaled['spotify_track_popularity'] = regression_data$spotify_track_popularity

regression_data_encoded_scaled
```

#### Regressions

<!-- 1.  split the dataset -->

<!-- ```{r} -->
<!-- set.seed(45) -->
<!-- split <- initial_split(head(regression_data_encoded,n_samples_to_consider), prop = 0.8) -->
<!-- train_data <- training(split) -->
<!-- test_data  <- testing(split) -->

<!-- # resampling for cross validation -->
<!-- cv_folds <- vfold_cv(train_data, v = 5) -->
<!-- ``` -->

<!-- 2.  define recipe, models and workflows -->

<!-- ```{r} -->
<!-- # recipe -->
<!-- recipe_spec <- recipe(spotify_track_popularity ~ ., data = train_data) |> -->
<!--   step_dummy(all_nominal_predictors(), one_hot = TRUE) -->

<!-- # models with tuning parameters -->
<!-- task = 'regression' -->
<!-- models <- list( -->
<!--   rf = rand_forest(mtry = tune(), trees = 500) |> -->
<!--     set_engine("randomForest") |> set_mode(task), -->

<!--   xgb = boost_tree(trees = 1000, tree_depth = tune(), learn_rate = tune()) |> -->
<!--     set_engine("xgboost") |> set_mode(task), -->

<!--   knn = nearest_neighbor(neighbors = tune()) |> -->
<!--     set_engine("kknn") |> set_mode(task) -->
<!-- ) -->

<!-- # workflows -->
<!-- workflows_list <- purrr::imap(models, function(model, name) { -->
<!--   workflow() |> add_model(model) |> add_recipe(recipe_spec) -->
<!-- }) -->
<!-- ``` -->

<!-- 3.  tuning -->

<!-- ```{r} -->
<!-- # tune_race_anova -->
<!-- tuned_models <- purrr::imap(workflows_list, function(wf, name) { -->
<!--   tune_race_anova( -->
<!--     wf, -->
<!--     resamples = cv_folds, -->
<!--     grid = 5, -->
<!--     metrics = metric_set(rmse), -->
<!--     control = control_race(verbose_elim = TRUE) -->
<!--   )}) -->

<!-- # extract best values -->
<!-- best_models <- purrr::map(tuned_models, ~ select_best(.x)) -->
<!-- ``` -->

<!-- 4.  finalize and fits the models to the training set -->

<!-- ```{r} -->
<!-- final_models <- purrr::map2(workflows_list, best_models, ~ finalize_workflow(.x, .y) |> fit(data = train_data)) -->
<!-- ``` -->

<!-- 5.  evaluation on the test set -->

<!-- ```{r} -->
<!-- results <- purrr::map(final_models, ~ { -->
<!--   predict(.x, test_data) |> -->
<!--     bind_cols(test_data) |> -->
<!--     metrics(truth = spotify_track_popularity, estimate = .pred)}) -->

<!-- # summary RMSE -->
<!-- results_df <- bind_rows(results, .id = "model") |> -->
<!--   filter(.metric == "rmse") |> arrange(.estimate) -->

<!-- print(results_df) -->
<!-- ``` -->

```{r big-model, cache=TRUE}
library(tidymodels)
library(furrr)
set.seed(45)

plan(multisession, workers = 6)

# 1. Split
split <- initial_split(regression_data_encoded_scaled, prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)

# 2. Recipe
recipe_spec <- recipe(spotify_track_popularity ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

# 3. Resampling (non usato direttamente, ma mantenuto)
cv_folds <- vfold_cv(train_data, v = 5)

# 4. Modelli con parametri fissi
models <- list(
  rf = rand_forest(mtry = 5, trees = 500) %>%
    set_engine("randomForest") %>% set_mode("regression"),

  xgb = boost_tree(trees = 1000, tree_depth = 6, learn_rate = 0.1) %>%
    set_engine("xgboost") %>% set_mode("regression"),

  knn = nearest_neighbor(neighbors = 7) %>%
    set_engine("kknn") %>% set_mode("regression")
)

# 5. Workflow per ciascun modello
workflows_list <- purrr::imap(models, function(model, name) {
  workflow() %>%
    add_model(model) %>%
    add_recipe(recipe_spec)
})

# 6. Fit parallelo
final_models <- furrr::future_map(workflows_list, function(wf) {
  library(tidymodels)  # necessario dentro il worker
  fit(wf, data = train_data)
})

# 7. Predizione + metriche parallele
results <- furrr::future_map(final_models, ~ {
  predict(.x, test_data) %>%
    bind_cols(test_data) %>%
    metrics(truth = spotify_track_popularity, estimate = .pred)
})

# 8. Riassunto risultati
results_df <- bind_rows(results, .id = "model") %>%
  filter(.metric == "rmse") %>%
  arrange(.estimate)

print(results_df)

# Torna a sequenziale
plan(sequential)
```

```{r}
all_predictions <- purrr::imap(final_models, ~ {
  predict(.x, test_data) |>
    bind_cols(test_data) |>
    mutate(model = .y)
}) |> bind_rows()

```

```{r}
ggplot(all_predictions, aes(x = spotify_track_popularity, y = .pred)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~model, scales = "free") +
  labs(
    title = "Predictions vs Real Values for Models",
    x = "Real Values (spotify_track_popularity)",
    y = "Predictions"
  ) +
  theme_minimal()
```

features importance

```{r}
# VIP plot list
vip_plots <- list()

# Random Forest
rf_fit <- extract_fit_parsnip(final_models$rf)
if (inherits(rf_fit$fit, "randomForest")) {
  vip_plots$rf <- vip(rf_fit$fit, num_features = 15) + ggtitle("Random Forest")
}

# XGBoost
xgb_fit <- extract_fit_parsnip(final_models$xgb)
if (inherits(xgb_fit$fit, "xgb.Booster")) {
  vip_plots$xgb <- vip(xgb_fit$fit, num_features = 15) + ggtitle("XGBoost")
}

# visualize plots
library(patchwork)

# show them
if (length(vip_plots) == 2) {
  vip_plots$rf + vip_plots$xgb
} else {
  # if only one is possible
  vip_plots[[1]]
}
```

### 4.2 Classification- predict which level of popularity

#### Preprocessing

Remove unrelevant features

```{r}
classification_data <- full_data_aggregate %>% select(-c( song_id, min_position, max_position, avg_position, weeks_in_chart, final_position))
print(head(classification_data,5))
```

Encode popularity

```{r}
classification_data <- encode_spotify_popularity(classification_data)

classification_data <- classification_data |>
  filter(!is.na(spotify_track_popularity)) 
```

Encode genre

```{r}
classification_data <- extract_first_genre(classification_data)
classification_data <- process_and_encode_genre(classification_data, top_n = 15)
```

```{r}
#print number of NaN for each column
#count_na_per_column_df(classification_data)
```

```{r}
print(head(classification_data,5))
```

Analyse the correlations

```{r}
# select only numerical data
numeric_data_clss <- classification_data |>
  select(where(is.numeric))

# compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# correlation heatmap
corrplot(cor_matrix, method = "color", 
         type = "upper", tl.col = "black", 
         #number.cex = 0.7, addCoef.col = "black",
         tl.cex = 0.8, col = colorRampPalette(c("purple", "white", "cyan"))(100))

#ggsave("4_regression_heatmap.png", width = 12, height = 16)
```

Target Distribution

```{r}
ggplot(classification_data, aes(x = spotify_track_popularity)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Popularity Distribution",
       x = "Spotify Track Popularity",
       y = "Counter") +
  theme_minimal()
```

split

```{r}
set.seed(45)
spotify_train_test <- classification_data |> initial_split(prop =  0.7)
spotify_train <- training(spotify_train_test) 
spotify_test  <- testing(spotify_train_test)
```

#### Classification

Random forest

```{r}
rand_forest_spec <- 
  rand_forest(mtry = 5, trees = 200) |>  
  set_engine('randomForest') |> 
  set_mode('classification')

rf_recipe = recipe(spotify_track_popularity~.,
                   data=spotify_train)

rf_wf <- workflow() |> 
  add_recipe(rf_recipe) |> 
  add_model(rand_forest_spec)

rf_wf_fit <- rf_wf |> 
  fit(spotify_train)

rf_wf_fit
```

```{r}
spotify_test_rf <- rf_wf_fit |> 
  augment(spotify_test)

# confusion matrix
cm_rF <- spotify_test_rf |> 
  conf_mat(truth = spotify_track_popularity,
           estimate = .pred_class)

# visualize
plot_conf_matrix(cm_rF, title = "Confusion Matrix - Classification Random Forest")

```

```{r}
head(summary(cm_rF),4)
```

```{r}
# select(spotify_test_rf, starts_with(".pred_"))

rf_wf_fit |> 
  extract_fit_engine() |> 
  vip()
```

<<<<<<<<<<<TODO>>>>>>>>>>>>>
> Is the music changed over time?
  
## 5. Taking only the last 25 years of data, <<<<<<<<the model will improve / possibile stimare meglio>>>>>>>>>>>>>?
### 5.1 Preprocessing
  Remove unrelevant features
```{r}
data_filtered <- full_data_clean |> select(-c(week_id, week_position, previous_week_position, peak_position, weeks_on_chart))
```

convert milliseconds in minutes
```{r}
data_filtered$spotify_track_duration_min <- data_filtered$spotify_track_duration_ms/60000
data_filtered <- data_filtered |> select(-c(spotify_track_duration_ms))
```

Filter the song after 2000
```{r}
bound_year = 2010

data_filtered_after <- data_filtered[data_filtered$year >= bound_year & data_filtered$year < 2021, ]
data_filtered_before <- data_filtered[data_filtered$year < bound_year, ]
```

encode the most frequent 15 genres for each period
```{r}
data_filtered_after_encoded <- extract_first_genre(data_filtered_after) 
data_filtered_after_encoded <- process_and_encode_genre(data_filtered_after_encoded, top_n = 15)

data_filtered_before_encoded <- extract_first_genre(data_filtered_before) 
data_filtered_before_encoded <- process_and_encode_genre(data_filtered_before_encoded, top_n = 15)
```

remove duplicates
```{r}
data_filtered_after_encoded$year <- NULL
data_filtered_after_encoded <- unique(data_filtered_after_encoded)
data_filtered_before_encoded <- unique(data_filtered_before_encoded)
```

```{r}
data_filtered_after_encoded
```

### 5.2 Regression after 2000
#### Preprocessing
```{r}
regression_data_after <- data_filtered_after_encoded |> select(-c(performer, song_id))
```

Scale the features
<<<<<<<<<<<<<<<<<TODO: vedere se mettere uno unique per ogni anno --> problema riassegnare target>>>>>>>>>>>>>>>>>>>>
```{r}
#regression_data_after <- robustHD::standardize(regression_data_after |> select(-c( year, spotify_track_popularity)))
regression_data_after$spotify_track_popularity <- data_filtered_after_encoded$spotify_track_popularity
regression_data_after <- unique(regression_data_after)

#regression_data_after$year <- data_filtered_after_encoded$ye
regression_data
```

#### Regressions
<!-- 1. split the dataset -->
<!-- ```{r} -->
<!-- registerDoParallel(cores = parallel::detectCores() - 1) -->

<!-- set.seed(45) -->
<!-- split <- initial_split(head(regression_data_after, n_samples_to_consider), prop = 0.8) -->
<!-- train_data <- training(split) -->
<!-- test_data  <- testing(split) -->

<!-- # resampling for cross validation -->
<!-- cv_folds <- vfold_cv(train_data, v = 5) -->
<!-- ``` -->

<!-- 2. define recipe, models and workflows -->
<!-- ```{r} -->
<!-- # recipe -->
<!-- recipe_spec <- recipe(spotify_track_popularity ~ ., data = train_data) |> -->
<!--   step_dummy(all_nominal_predictors(), one_hot = TRUE) -->

<!-- # models with tuning parameters -->
<!-- task = 'regression' -->
<!-- models <- list( -->
<!--   rf = rand_forest(mtry = tune(), trees = 500) |> -->
<!--     set_engine("randomForest") |> set_mode(task), -->

<!--   xgb = boost_tree(trees = 1000, tree_depth = tune(), learn_rate = tune()) |> -->
<!--     set_engine("xgboost") |> set_mode(task), -->

<!--   knn = nearest_neighbor(neighbors = tune()) |> -->
<!--     set_engine("kknn") |> set_mode(task) -->
<!-- ) -->

<!-- # workflows -->
<!-- workflows_list <- purrr::imap(models, function(model, name) { -->
<!--   workflow() |> add_model(model) |> add_recipe(recipe_spec) -->
<!-- }) -->
<!-- ``` -->

<!-- 3. tuning -->
<!-- ```{r} -->
<!-- # tune_race_anova -->
<!-- tuned_models <- purrr::imap(workflows_list, function(wf, name) { -->
<!--   tune_race_anova( -->
<!--     wf, -->
<!--     resamples = cv_folds, -->
<!--     grid = 5, -->
<!--     metrics = metric_set(rmse), -->
<!--     control = control_race(verbose_elim = TRUE) -->
<!--   )}) -->

<!-- # extract best values -->
<!-- best_models <- purrr::map(tuned_models, ~ select_best(.x)) -->
<!-- ``` -->

<!-- 4. finalize and fits the models to the training set -->
<!-- ```{r} -->
<!-- final_models <- purrr::map2(workflows_list, best_models, ~ finalize_workflow(.x, .y) |> fit(data = train_data)) -->
<!-- ``` -->

<!-- 5. evaluation on the test set -->
<!-- ```{r} -->
<!-- results <- purrr::map(final_models, ~ { -->
<!--   predict(.x, test_data) |> -->
<!--     bind_cols(test_data) |> -->
<!--     metrics(truth = spotify_track_popularity, estimate = .pred)}) -->

<!-- # summary RMSE -->
<!-- results_df <- bind_rows(results, .id = "model") |> -->
<!--   filter(.metric == "rmse") |> arrange(.estimate) -->

<!-- print(results_df) -->
<!-- ``` -->

```{r}
# compute correlation matrix
cor_matrix <- cor(regression_data_after, use = "complete.obs")

# correlation heatmap
corrplot(cor_matrix, method = "color", 
         type = "upper", tl.col = "black", 
         #number.cex = 0.7, addCoef.col = "black",
         tl.cex = 0.8, col = colorRampPalette(c("purple", "white", "cyan"))(100))
```



```{r small-model, cache=True}
set.seed(45)
library(tidymodels)
library(furrr)

plan(multisession, workers = 6)

# 1. Split
split <- initial_split(regression_data_after, prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)

# 2. Recipe
recipe_spec <- recipe(spotify_track_popularity ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

# 3. Resampling (non usato qui)
cv_folds <- vfold_cv(train_data, v = 5)

# 4. Modelli con parametri fissi
models <- list(
  rf = rand_forest(mtry = 5, trees = 500) %>%
    set_engine("randomForest") %>% set_mode("regression"),

  xgb = boost_tree(trees = 1000, tree_depth = 6, learn_rate = 0.1) %>%
    set_engine("xgboost") %>% set_mode("regression"),

  knn = nearest_neighbor(neighbors = 7) %>%
    set_engine("kknn") %>% set_mode("regression")
)

# 5. Workflows
workflows_list <- purrr::imap(models, function(model, name) {
  workflow() %>%
    add_model(model) %>%
    add_recipe(recipe_spec)
})

# 6. Fit parallelo
final_models <- furrr::future_map(workflows_list, function(wf) {
  library(tidymodels)  # necessario dentro il worker
  fit(wf, data = train_data)
})

# 7. Valuta su test set in parallelo
results <- furrr::future_map(final_models, ~ {
  predict(.x, test_data) %>%
    bind_cols(test_data) %>%
    metrics(truth = spotify_track_popularity, estimate = .pred)
})

# 8. Riassunto RMSE
results_df <- bind_rows(results, .id = "model") %>%
  filter(.metric == "rmse") %>%
  arrange(.estimate)

print(results_df)

# (opzionale) chiudi il cluster
plan(sequential)
```
```{r}
regression_data_after
```


```{r}
all_predictions <- purrr::imap(final_models, ~ {
  predict(.x, test_data) |>
    bind_cols(test_data) |>
    mutate(model = .y)
}) |> bind_rows()

```

```{r}
ggplot(all_predictions, aes(x = spotify_track_popularity, y = .pred)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~model, scales = "free") +
  labs(
    title = "Predictions vs Real Values for Models",
    x = "Real Values (spotify_track_popularity)",
    y = "Predictions"
  ) +
  theme_minimal()
```

features importance
```{r}
# VIP plot list
vip_plots <- list()

# Random Forest
rf_fit <- extract_fit_parsnip(final_models$rf)
if (inherits(rf_fit$fit, "randomForest")) {
  vip_plots$rf <- vip(rf_fit$fit, num_features = 15) + ggtitle("Random Forest")
}

# XGBoost
xgb_fit <- extract_fit_parsnip(final_models$xgb)
if (inherits(xgb_fit$fit, "xgb.Booster")) {
  vip_plots$xgb <- vip(xgb_fit$fit, num_features = 15) + ggtitle("XGBoost")
}

# visualize plots
library(patchwork)

# show them
if (length(vip_plots) == 2) {
  vip_plots$rf + vip_plots$xgb
} else {
  # if only one is possible
  vip_plots[[1]]
}
```
```{r}

```

QQ plot
```{r}
numeric_data <- regression_data |> select(where(is.numeric))
par(mfrow = c(2,2))  # cambia a seconda del numero di variabili

for (colname in names(numeric_data)) {
  qqnorm(numeric_data[[colname]], main = paste("QQ Plot -", colname))
  qqline(numeric_data[[colname]])
}
```

> Sono questi due periodi diversi?? Guardiamo i primi due elementi

```{r}
```


```{r}
loudness_summary <- full_data_clean %>%
  group_by(year) %>%
  summarise(
    mean_loudness = mean(loudness, na.rm = TRUE),
    sd_loudness = sd(loudness, na.rm = TRUE),
    n = n(),
    se = sd_loudness / sqrt(n)  # standard error
  )

ggplot(loudness_summary, aes(x = year, y = mean_loudness)) +
  geom_line(color = "steelblue", size = 1) +
  geom_ribbon(aes(ymin = mean_loudness - se, ymax = mean_loudness + se), 
              fill = "lightblue", alpha = 0.4) +
  labs(title = "Loudness Through the Years",
       x = "Year", y = "Average Loudness (dB)") +
  theme_minimal()
```


```{r}

loudness_summary <- full_data_clean %>%
  group_by(year) %>%
  summarise(
    mean_loudness = mean(spotify_track_duration_ms, na.rm = TRUE),
    sd_loudness = sd(spotify_track_duration_ms, na.rm = TRUE),
    n = n(),
    se = sd_loudness / sqrt(n)  # standard error
  )

ggplot(loudness_summary, aes(x = year, y = mean_loudness)) +
  geom_line(color = "steelblue", size = 1) +
  geom_ribbon(aes(ymin = mean_loudness - se, ymax = mean_loudness + se), 
              fill = "lightblue", alpha = 0.4) +
  labs(title = "Track duration over the Years",
       x = "Year", y = "track duration (ms)") +
  theme_minimal()

```
```{r}

loudness_summary <- full_data_clean %>%
  group_by(year) %>%
  summarise(
    mean_loudness = mean(acousticness, na.rm = TRUE),
    sd_loudness = sd(acousticness, na.rm = TRUE),
    n = n(),
    se = sd_loudness / sqrt(n)  # standard error
  )

ggplot(loudness_summary, aes(x = year, y = mean_loudness)) +
  geom_line(color = "steelblue", size = 1) +
  geom_ribbon(aes(ymin = mean_loudness - se, ymax = mean_loudness + se), 
              fill = "lightblue", alpha = 0.4) +
  labs(title = "Acousticness over the Years",
       x = "Year", y = "Acousticness") +
  theme_minimal()

```


## 6. Have songs from the last ten years become louder than before? We chose to perform this hypothesis test on the loudness because we noticed it was the most important feature among all the audio features of the songs.

### 6.1 Hypothesis test
#### Preprocessing

Remove unrelevant features
```{r}
# selecting only the numeric features
hypothesis_data_after <- data_filtered_after_encoded |> select(where(is.numeric))
hypothesis_data_before <- data_filtered_before_encoded |> select(where(is.numeric))
```

Analyse distribution
```{r}
#selecting all the features that are not related to the genre
numeric_data_after_no_genre <- hypothesis_data_after |>
  select(1:(ncol(hypothesis_data_after) - 11))
par(mfrow = c(2,2))

numeric_data_before_no_genre <- hypothesis_data_before |>
  select(1:(ncol(hypothesis_data_before) - 11))
```

*After 2000*
```{r}
#plotting the QQ plots to show the distribution of the dataset
par(mfrow = c(2,2))
for (colname in names(numeric_data_after_no_genre)) {
  qqnorm(numeric_data_after_no_genre[[colname]], main = paste("QQ Plot -", colname))
  qqline(numeric_data_after_no_genre[[colname]])
}
```
*Before 2000*
```{r}
#plotting the QQ plots to show the distribution of the dataset
par(mfrow = c(2,2))
for (colname in names(numeric_data_before_no_genre)) {
  qqnorm(numeric_data_before_no_genre[[colname]], main = paste("QQ Plot -", colname))
  qqline(numeric_data_before_no_genre[[colname]])
}
```

Based on the QQ plots for the periods before and after 2015, none of the feature distributions appear to be normal. Therefore, to compare the two distributions, we should use the Wilcoxon test.

#### Hypothesis Test
```{r}
#Doing a wilcox test based on the loudness of the songs
wilcox.test(
  hypothesis_data_after$spotify_track_duration_min,
  hypothesis_data_before$spotify_track_duration_min,
  alternative = "less"
)

#filtering the datasets with only the loudness feature
loudness_after <- hypothesis_data_after$spotify_track_duration_min
loudness_before <- hypothesis_data_before$spotify_track_duration_min


plot(density(loudness_after), col = "blue", lwd = 2,
     main = "Spotify track duration min Density: After 2010 vs Before 2010", xlab = "duration")
lines(density(loudness_before), col = "red", lwd = 2)

legend("topright", legend = c("After 2010", "Before 2010"), col = c("blue", "red"), lwd = 2)
```

As we can see from the two distributions, songs released after 2015 are louder than those released before. Why is that? This could be due to the evolution of music over the years—listeners' tastes have changed, and artists have explored new styles, adapting their sound to the cultural and historical context.


## Question 2 - As a manager, should I continue to invest in this song given its statistics? Will it remain popular over time?

### Regression - predict the position on the Billboard

#### Preprocessing

Remove unrelevant features

```{r}
regression_data <- full_data_aggregate %>% select(-c(song_id))
print(head(regression_data,5))
```

one hot encoding of the most 15 important genres

```{r}
regression_data <- extract_first_genre(regression_data)
regression_data_encoded <- process_and_encode_genre(regression_data, top_n = 15)
```

```{r}
print(head(regression_data_encoded, 5))
```

#### Regression

```{r}
set.seed(45)
split <- initial_split(head(regression_data_encoded,100), prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)

# recipe
recipe_spec <- recipe(final_position ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

task = 'regression'

# models
models <- list(
  rf = rand_forest(mtry = 5) %>% set_engine("randomForest") %>% set_mode(task),
  #lm = linear_reg() %>% set_engine("lm") %>% set_mode(task),
  xgb = boost_tree(trees = 100, tree_depth = 10) %>% set_engine("xgboost") %>% set_mode(task),
  knn = nearest_neighbor(neighbors = 5) %>% set_engine("kknn") %>% set_mode(task)
)

# workflow
workflows_list <- purrr::map(models, ~ workflow() %>% add_model(.x) %>% add_recipe(recipe_spec))

# train models
fitted_models <- purrr::map(workflows_list, ~ fit(.x, data = train_data))

# evaluation
results <- purrr::map(fitted_models, ~ {
  predict(.x, test_data) %>%
    bind_cols(test_data) %>%
    metrics(truth = final_position, estimate = .pred)
})

# results
results_df <- bind_rows(results, .id = "model")

results_df %>%
  filter(.metric == "rmse") %>%
  arrange(.estimate)

print(results_df)
```

```{r}
results <- list()

for (model_name in names(fitted_models)) {
  model <- models[[model_name]]
  
  wf <- workflow() %>%
    add_model(model) %>%
    add_recipe(recipe_spec)
  
  fit_model <- fit(wf, data = train_data)
  
  preds <- predict(fit_model, test_data) %>%
    bind_cols(test_data) %>%
    mutate(model = model_name)
  
  results[[model_name]] <- preds
}

all_predictions <- bind_rows(results)

```

```{r}
ggplot(all_predictions, aes(x = final_position, y = .pred)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~model, scales = "free") +
  labs(
    title = "Predictions vs Real Values",
    x = "Real Values (spotify_track_popularity)",
    y = "Predictions"
  ) +
  theme_minimal()
```

```{r}
# VIP plot list
vip_plots <- list()

# Random Forest
rf_fit <- extract_fit_parsnip(fitted_models$rf)
if (inherits(rf_fit$fit, "randomForest")) {
  vip_plots$rf <- vip(rf_fit$fit, num_features = 15) + ggtitle("Random Forest - Variable Importance")
}

# XGBoost
xgb_fit <- extract_fit_parsnip(fitted_models$xgb)
if (inherits(xgb_fit$fit, "xgb.Booster")) {
  vip_plots$xgb <- vip(xgb_fit$fit, num_features = 15) + ggtitle("XGBoost - Variable Importance")
}

# visualize plots
library(patchwork)

# show them
if (length(vip_plots) == 2) {
  vip_plots$rf + vip_plots$xgb
} else {
  # if only one is possible
  vip_plots[[1]]
}
```

### Classification- predict whether a sign will be in the top 30

#### Preprocessing

```{r}
classification_data <- full_data_aggregate |> select(-c( song_id))
print(head(classification_data,5))
```

Encode popularity

```{r}
classification_data <- encode_spotify_popularity(classification_data)

classification_data <- classification_data |>
  filter(!is.na(spotify_track_popularity)) 
```

Encode genre

```{r}
classification_data <- extract_first_genre(classification_data)
classification_data <- process_and_encode_genre(classification_data, top_n = 5)
```

```{r}
#print number of NaN for each column
count_na_per_column_df(classification_data)
```

```{r}
#create the new variable top30: whether a song is in the top 30
classification_data <- classification_data |>
  mutate(top30 = ifelse(final_position <= 30, 1, 0))
# remove final_position
classification_data <- classification_data |> select(-c( final_position))
print(head(classification_data,5))
```

```{r}
# select only numerical data
numeric_data_clss <- classification_data |>
  select(where(is.numeric))

# correlation matrix
cor_matrix <- cor(numeric_data_clss, use = "complete.obs")

ggcorrplot(cor_matrix, 
           method = "square",     # or "circle"
           type = "lower",        # only lower triangle        
           colors = c("red", "white", "blue"),
           title = "Correlation Matrix",
           ggtheme = ggplot2::theme_minimal())
```

```{r}
hist(classification_data$top30)
```

#### Classification

```{r}
set.seed(42)
classification_data$top30 <- as.factor(classification_data$top30)
spotify_train_test <- classification_data |> initial_split(prop =  0.7)
spotify_train <- training(spotify_train_test) 
spotify_test  <- testing(spotify_train_test)
```

```{r}
rand_forest_spec <- 
  rand_forest(mtry = 5, trees = 200) |>  
  set_engine('randomForest') |> 
  set_mode('classification')

rf_recipe = recipe(top30~.,
                   data=spotify_train)

rf_wf <- workflow() |> 
  add_recipe(rf_recipe) |> 
  add_model(rand_forest_spec)

rf_wf_fit <- rf_wf |> 
  fit(spotify_train)

rf_wf_fit
```

```{r}
spotify_test_rf <- rf_wf_fit |> 
  augment(spotify_test)

# confusion matrix
cm_rF <- spotify_test_rf |> 
  conf_mat(truth = top30,
           estimate = .pred_class)

# visualize
plot_conf_matrix(cm_rF, title = "Confusion Matrix - Classification Random Forest")
```

```{r}
head(summary(cm_rF),4)
```

```{r}
roc_rF<-roc_curve(spotify_test_rf,
          top30,
          .pred_0,event_level = "first")
roc_rF |> autoplot()
```

```{r}
rf_wf_fit |> 
  extract_fit_engine() |> 
  vip()
```


