---
title: "R Notebook"
output: html_notebook
---

# Spotify Project

Authors: [Bernacchia Alessia](https://github.com/AlessiaBernacchia), [Pioda Tommaso](https://github.com/Thetommigun432), [Villani Giacomo](https://github.com/DownToTheGround)

```{r}
library(vip)
library(readr)
library(dplyr)
library(tidymodels)
library(ggplot2)
library(tidymodels)
library(rsample)
library(tidyr)
library(themis)
library(stringr)
library(purrr)
library(caret)
library(kknn)
library(ggcorrplot)
#install.packages("xgboost")
library(parsnip)
library(randomForest)
library(kknn)
library(xgboost)
library(finetune)
library(doParallel)
library(robustHD)
library(parsnip)
```

*Functions*:
```{r}
# count NaN for each column
count_na_per_column <- function(data) {
  sapply(data, function(col) sum(is.na(col)))}

count_na_per_column_df <- function(data) {
  na_counts <- sapply(data, function(col) sum(is.na(col)))
  na_counts_df <- data.frame(
    feature = names(na_counts),
    na_count = na_counts
  )
  na_counts_df %>% arrange(desc(na_count))
}

```


## Import the data

```{r}
billboard <- read_csv("billboard.csv", 
                           col_types = cols(
                            url = col_skip(), 
                            song = col_skip(),
                            instance = col_skip()
                            ))
print(head(billboard,5))
```

```{r}
audio_features <- read_csv("audio_features.csv", 
                          col_types = cols( 
                            performer = col_skip(), 
                            song = col_skip(), 
                            spotify_track_preview_url = col_skip(),
                            spotify_track_explicit = col_skip(),
                            spotify_track_album = col_skip(),
                            spotify_track_id = col_skip()
                            ))
print(head(audio_features,5))
```

```{r}
billboard <- na.omit(billboard) # remove all NaN
audio_features <- as_tibble(audio_features) # translate in a modern and more flexible version
audio_features <- na.omit(audio_features) # remove all NaN

```

```{r}
full_data <- left_join(billboard, audio_features, by = "song_id") # merge datasets
```

## Data exploration and cleaning
```{r}
glimpse(full_data)
```

### Missing Values

```{r}
na_counts <- count_na_per_column_df(full_data)
print(na_counts)
```


```{r}
full_data_clean <- full_data
full_data_clean$previous_week_position[is.na(full_data_clean$previous_week_position)] <- 0 # substitute NaN with 0
full_data_clean <- na.omit(full_data_clean) # remove all NaN

nrow(full_data_clean)
ncol(full_data_clean) 

```
### Distributions
Distribution of numeric data

```{r}
numeric_data <- full_data_clean %>% select(where(is.numeric))

numeric_data_long <- numeric_data %>% pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# histograms distribution facet
ggplot(numeric_data_long, aes(x = value)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 25) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +  # 3 columns
  theme_minimal()
#ggsave("0_numeric_histogram_distribution.png", width = 12, height = 16)
```

Distribution of categorical data
(due the fact there are too many unique values, we choose to filter and observe only the first 50 most frequent variables)
```{r}
categorical_data <- full_data_clean %>% select(where(~ !is.numeric(.)))

categorical_data_long <- categorical_data %>% pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# select top 50
categorical_data_top50 <- categorical_data_long %>%
  group_by(variable, value) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(variable) %>%
  slice_max(order_by = count, n = 50)

# barplot distribution facet
ggplot(categorical_data_top50, aes(x = value, y = count)) +
  geom_col(fill = "steelblue", color = "white") + 
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  theme_minimal() +
  theme(#axis.text.x = element_text(angle = 45, hjust = 1))
    axis.text.x = element_blank() )
#ggsave("0_categorical_histogram_distribution.png", width = 12, height = 16)
```
### Cleaning
```{r}
hist(full_data_clean$instrumentalness)
```

```{r}
hist(full_data_clean$time_signature)
```

```{r}
hist(full_data_clean$speechiness)
```


```{r}
hist(full_data_clean$spotify_track_duration_ms)
```
From the distribution of `instrumentalness` and `time signature` they seem to be not significant for our purpose. Then we decide to remove them.
```{r}
full_data_clean <- full_data_clean %>% select(-c(instrumentalness, time_signature))
```
### Outliers
```{r}
# histograms distribution facet
ggplot(numeric_data_long, aes(x = value, y = "")) +
  geom_boxplot(fill = "steelblue") +
  facet_wrap(~ variable, scales = "free", ncol = 3) + 
  theme_minimal()
#ggsave("0_numeric_boxplot_distribution.png", width = 12, height = 16)
```
```{r}
# correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

ggcorrplot(cor_matrix, 
           method = "square",      # o "circle"
           type = "lower",         # solo triangolo inferiore         
           colors = c("red", "white", "blue"),  # come in seaborn
           title = "Matrice di Correlazione",
           ggtheme = ggplot2::theme_minimal())
```


```{r}
# new features creation
billboard_summary <- billboard %>%
  group_by(song_id) %>%
  summarise(
    min_position = min(week_position),
    max_position = max(week_position),
    avg_position = mean(week_position),
    weeks_in_chart = n_distinct(week_id),
    final_position = last(week_position)
  )

full_data_aggregate <- merge(billboard_summary, audio_features, by = "song_id", all = FALSE)
full_data_aggregate <- full_data_aggregate %>% select(-c(instrumentalness, time_signature))

```

```{r}
print(head(full_data_aggregate, 5))
```


```{r}
nrow(billboard)
nrow(billboard_summary)
nrow(audio_features)
nrow(full_data)
nrow(full_data_aggregate)
```


## Regression - predict the position on the billboard

### Preprocessing
Remove unrelevant features

```{r}
regression_data <- full_data_aggregate %>% select(-c(song_id, min_position, max_position, avg_position, weeks_in_chart, final_position))
print(head(regression_data,5))
```


```{r}
# removes useless not required characters
regression_data$spotify_genre <- gsub("\\[|\\]|'", "", regression_data$spotify_genre)
# create a list, with div = ,
regression_data$spotify_genre <- strsplit(regression_data$spotify_genre, split = ",")
# extract only the first genre for each line
regression_data$first_genre <- sapply(regression_data$spotify_genre, function(x) x[1])

regression_data$first_genre <- sapply(regression_data$spotify_genre, function(x) {
  if (length(x) == 0) {
    return("others")
  } else {
    return(x[1])
  }
})

a='audio_features$spotify_track_popularity <- sapply(audio_features$spotify_track_popularity, function(x) {
  if (x <= 25) {
    return("low")
  }
  if (x > 25 & x <= 50) {
    return("medium")
  }
  if (x > 50 & x <= 75) {
    return("high")
  }
    if (x > 75 & x <= 100) {
    return("very high")
  } })'
```
select only important genres (>=200 density)
```{r}
# count frequency of each sample in first_genre column
genre_counts <- table(regression_data$first_genre)

# identify genres with less than 200 samples
low_count_genres <- names(genre_counts[genre_counts < 300])

# substitute those genres with "others"
regression_data$first_genre <- ifelse(regression_data$first_genre %in% low_count_genres, "others", regression_data$first_genre)

# verify the counts after grouping
genre_counts_after_grouping <- table(regression_data$first_genre)

# drop original spotify_genre column
regression_data <- subset(regression_data, select = -c(spotify_genre))

# ensure the column is upgraded for dummyVars
df <- data.frame(first_genre = regression_data$first_genre)

# create dummy variables (one-hot encoding)
dummy <- dummyVars(~ first_genre, data = df)
final_df <- data.frame(predict(dummy, newdata = df))

# bind dummy variables to the rest of regression_data
regression_data_encoded_1 <- bind_cols(
  regression_data %>% select(-first_genre),
  final_df
)
```


```{r}
# frequency counter
genre_counts <- table(regression_data$first_genre)

# order based in descrescent order the frequency
genre_counts_sorted <- sort(genre_counts, decreasing = TRUE)

# take only first 15 genres
top_15_genres <- names(genre_counts_sorted)[1:15]

# named others as "others"
regression_data$first_genre <- ifelse(
  regression_data$first_genre %in% top_15_genres,
  regression_data$first_genre,
  "others"
)

# verify
genre_counts_after_grouping <- table(regression_data$first_genre)

# drop original spotify_genre column
#regression_data <- subset(regression_data, select = -c(spotify_genre))

# ensure the column is upgraded for dummyVars
df <- data.frame(first_genre = regression_data$first_genre)

# create dummy variables (one-hot encoding)
dummy <- dummyVars(~ first_genre, data = df)
final_df <- data.frame(predict(dummy, newdata = df))

# bind dummy variables to the rest of regression_data (without first_genre)
regression_data_encoded <- bind_cols(
  regression_data %>% select(-first_genre),
  final_df
)

```
### Regression
```{r}
regression_data
```


```{r}
registerDoParallel(cores = parallel::detectCores() - 1)

set.seed(45)

regression_data_encoded_scaled <- robustHD::standardize(regression_data_encoded %>% select(-spotify_track_popularity))

regression_data_encoded_scaled['spotify_track_popularity'] = regression_data$spotify_track_popularity
```

```{r}
# 1. Split
split <- initial_split(head(regression_data_encoded_scaled,10000), prop = 0.8)
#split <- initial_split(regression_data_encoded_scaled, prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)

# 2. Recipe
recipe_spec <- recipe(spotify_track_popularity ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

# 3. Resampling
cv_folds <- vfold_cv(train_data, v = 5)

# 4. Definisci i modelli con parametri da tunare
models <- list(
  rf = rand_forest(mtry = tune(), trees = 500) %>%
    set_engine("randomForest") %>% set_mode("regression"),

  xgb = boost_tree(trees = 1000, tree_depth = tune(), learn_rate = tune()) %>%
    set_engine("xgboost") %>% set_mode("regression"),

  knn = nearest_neighbor(neighbors = tune()) %>%
    set_engine("kknn") %>% set_mode("regression")
)

# 5. Crea i workflow
workflows_list <- purrr::imap(models, function(model, name) {
  workflow() %>%
    add_model(model) %>%
    add_recipe(recipe_spec)
})

# 6. Tuning veloce con tune_race_anova
tuned_models <- purrr::imap(workflows_list, function(wf, name) {
  tune_race_anova(
    wf,
    resamples = cv_folds,
    grid = 5,
    metrics = metric_set(rmse),
    control = control_race(verbose_elim = TRUE)
  )
})

# 7. Estrai i migliori iperparametri
best_models <- purrr::map(tuned_models, ~ select_best(.x))

# 8. Finalizza e fitta
final_models <- purrr::map2(workflows_list, best_models, ~ finalize_workflow(.x, .y) %>% fit(data = train_data))

# 9. Valuta su test set
results <- purrr::map(final_models, ~ {
  predict(.x, test_data) %>%
    bind_cols(test_data) %>%
    metrics(truth = spotify_track_popularity, estimate = .pred)
})

# 10. Riassunto risultati RMSE
results_df <- bind_rows(results, .id = "model") %>%
  filter(.metric == "rmse") %>%
  arrange(.estimate)

print(results_df)
```

```{r}
all_predictions <- purrr::imap(final_models, ~ {
  predict(.x, test_data) %>%
    bind_cols(test_data) %>%
    mutate(model = .y)
}) %>% bind_rows()

```

```{r}
ggplot(all_predictions, aes(x = spotify_track_popularity, y = .pred)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~model, scales = "free") +
  labs(
    title = "Predizioni vs Valori Reali per Modello",
    x = "Valori Reali (spotify_track_popularity)",
    y = "Predizioni"
  ) +
  theme_minimal()
```

```{r}
vip(fitted_models)
```
```{r}
if (inherits(fitted_models$rf$fit, "randomForest")) {
  vip(fitted_models$rf$fit)
}
```


```{r}
# Filter and group data for "rock" genre
rock_loudness <- regression_data %>%
  filter(first_genre == 'album rock') %>%
  pull(loudness)

jazz_loudness <- regression_data %>%
  filter(first_genre == 'dance pop') %>%
  pull(loudness)


# Plot density for the "rock" genre
plot(density(rock_loudness), col = "blue", main = "Compare Densities")

# Assuming y is already defined as another dataset for comparison
lines(density(jazz_loudness), col = "red")

# Add legend
legend("topright", legend = c("Rock", "Y"), col = c("blue", "red"), lty = 1)

ks.test(rock_loudness, jazz_loudness)
```

```{r}
numeric_data <- regression_data |> select(where(is.numeric))
par(mfrow = c(2,2))  # cambia a seconda del numero di variabili

for (colname in names(numeric_data)) {
  qqnorm(numeric_data[[colname]], main = paste("QQ Plot -", colname))
  qqline(numeric_data[[colname]])
}
```


